---
title: 'One-sample hypothesis tests'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, rladies, rladies-fonts, "my-theme.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
--- 

```{r, echo = F, results = 'hide', message = F, warning = F}
library(here)
library(tidyverse)
library(psych)
library(knitr)
```


## Today...

* The chi-square goodness-of-fit test
* One-sample t-tests

---

### Key questions:

* How do we know if category frequencies are consistent with null hypothesis expectations?

* How do we handle categories with very low frequencies?

* How do we compare one sample to a population mean?

---

One-sample tests compare your given sample with a "known" population.

* Research question: does this sample come from this population?

**Hypotheses**

$H_0$: Yes, this sample comes from this population.

$H_1$: No, this sample comes from a different population. 



---

The sample data were obtained from Census at School, a website developed by the American Statistical Association to help students in the 4th through 12th grades understand statistical problem-solving. 

  * The site sponsors a survey that students can complete and a database that students and instructors can use to illustrate principles in quantitative methods.  
  * The database includes students from all 50 states, from grade levels 4 through 12, both boys and girls,  who have completed the survey dating back to 2010.  

---

We’ll focus on this one:

Which of the following superpowers would you most like to have? Select one.

* Invisibility
* Telepathy (read minds)
* Freeze time
* Super strength
* Fly

The responses from 247 randomly selected Oregon students were obtained from the Census at School database. 

---

```{r}
school = read_csv(here("data/census_at_school.csv"))
```

---

```{r frequency table, results = 'asis'}
school %>%
  group_by(Superpower) %>%
  summarize(Frequency = n()) %>%
  mutate(Proportion = Frequency/sum(Frequency)) %>%
  kable(., format = "html", digits = 2)
```

Descriptively this is interesting.  But, are the responses unusual or atypical in any way?  To answer that question, we need some basis for comparison—a null hypothesis.  One option would be to ask if the Oregon preferences are different compared to students from the general population.

---

class: center

```{r, echo = FALSE, message=FALSE, warning = F}
school_usa = read_csv(here("data/census_at_school_usa.csv"))

school %>%
  full_join(select(school_usa, Region, Superpower)) %>%
  mutate(Region = ifelse(Region == "OR", "OR", "USA")) %>%
  filter(!is.na(Superpower)) %>%
  group_by(Region, Superpower) %>%
  summarize(Frequency = n()) %>%
  mutate(Proportion = Frequency/sum(Frequency)) %>%
  ggplot(aes(x = Superpower, y = Proportion, fill = Region)) +
  geom_bar(stat = "identity", position = "dodge") +
  ggtitle("Category Proportion as a \nfunction of Source") +
  theme_bw(base_size = 20) + theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, vjust = .5))
```

---

$H_0$: Oregon student superpower preferences are similar to the preferences of typical students in the United States.

$H_1$: Oregon student superpower preferences are different from the preferences of typical students in the United States. 

```{r, echo = F, results = 'asis', message=F}
school %>%
  full_join(select(school_usa, Region, Superpower)) %>%
  mutate(Region = ifelse(Region == "OR", "OR", "USA")) %>%
  filter(!is.na(Superpower)) %>%
  group_by(Region, Superpower) %>%
  summarize(Frequency = n()) %>%
  mutate(Proportion = Frequency/sum(Frequency)) %>%
  select(-Frequency) %>%
  spread(Region, Proportion) %>%
  kable(., col.names = c("Superpower", "OR Observed Proportion", "USA Proportion"), format = "html", digits = 2)
```

---

To compare the Oregon observed frequencies to the US data, we need to calculate the frequencies that would have been expected if Oregon was just like all of the other states.

The expected frequencies under this null model can be obtained by taking each preference category proportion from the US data (the null expectation) and multiplying it by the sample size for Oregon:

$$E_i = P_iN_{OR}$$
---
```{r, echo = F}
usa_freq = table(school_usa$Superpower)
usa_prop = usa_freq/sum(usa_freq)

school %>%
  filter(!is.na(Superpower)) %>%
  group_by(Superpower) %>%
  summarize(Frequency = n()) %>%
  mutate(Expected = usa_prop*247) %>%
  kable(., format = "html", digits = 2, col.names = c("Superpower", "Observed\nFreq", "Expected Freq"))
```

Now what?  We need some way to index differences between these frequencies, preferably one that translates easily into a sampling distribution so that we can sensibly determine how rare or unusual the Oregon data are compared to the US (null) distribution.

---

$$\chi^2_{df = k-1} = \sum^k_{i=1}\frac{(O_i-E_i)^2}{E_i}$$

The chi-square goodness of fit (GOF) statistic compares observed and expected frequencies.  It is small when the observed frequencies closely match the expected frequencies under the null hypothesis.  The chi-square distribution can be used to determine the particular $\chi^2$ value that corresponds to a rare or unusual profile of observed frequencies.

---

```{r create obs, echo = 3:4}
or_observed = table(school$Superpower)
or_expected = (table(school_usa$Superpower)/sum(table(school_usa$Superpower)))*247
or_observed
or_expected
```

---

```{r, ref.label="create obs", echo=3:4}

```

```{r}
(chi_square = sum((or_observed - or_expected)^2/or_expected))
```

---

```{r, ref.label="create obs", echo=3:4}

```

```{r, highlight=2}
(chi_square = sum((or_observed - or_expected)^2/or_expected))
(critical_val = qchisq(p = 0.95, df = length(or_expected)-1))
```

---

```{r, ref.label="create obs", echo=3:4}

```

```{r, highlight=2}
(chi_square = sum((or_observed - or_expected)^2/or_expected))
(critical_val = qchisq(p = 0.95, df = length(or_expected)-1))
(p_val = pchisq(q = chi_square, df = length(or_expected)-1, lower.tail = F))
```

---

.left-column[
.small[
The degrees of freedom are the number of categories (k) minus 1.  Given that the category frequencies must sum to the total sample size, k-1 category frequencies are free to vary; the last is determined.

]
]

```{r, echo = FALSE}
data.frame(x = seq(0,20)) %>%
  ggplot(aes(x = x)) +
  stat_function(fun = function(x) dchisq(x, df = length(or_expected)-1), geom = "line") +
  stat_function(fun = function(x) dchisq(x, df = length(or_expected)-1), geom = "area", fill = "purple", 
                xlim =c(critical_val, 20)) +
  geom_vline(aes(xintercept = critical_val), linetype = 2, color = "purple")+
    geom_vline(aes(xintercept = chi_square), linetype = 2, color = "black")+
  geom_text(aes(x = critical_val+2, y = dchisq(critical_val, length(or_expected)-1) + .05, 
                label = paste("Critical Value =", round (critical_val,2))), angle = 90)+
    geom_text(aes(x = chi_square+2, y = dchisq(critical_val, length(or_expected)-1) + .05, 
                label = paste("Critical Value =", round (chi_square,2))), angle = 90)+
  labs(x = "Chi-Square", y = "Density", title = "Area under the curve") +
  theme_light(base_size = 20)
```

---
.left-column[
.small[
The degrees of freedom are the number of categories (k) minus 1.  Given that the category frequencies must sum to the total sample size, k-1 category frequencies are free to vary; the last is determined.

]
]

```{r, echo = FALSE}
data.frame(x = seq(0,20)) %>%
  ggplot(aes(x = x)) +
  stat_function(fun = function(x) dchisq(x, df = length(or_expected)-1), geom = "line") +
  stat_function(fun = function(x) dchisq(x, df = length(or_expected)-1), geom = "area", fill = "purple", 
                xlim =c(chi_square, 20)) +
  geom_vline(aes(xintercept = critical_val), linetype = 2, color = "black")+
  geom_vline(aes(xintercept = chi_square), linetype = 2, color = "purple")+
  geom_text(aes(x = critical_val+2, y = dchisq(critical_val, length(or_expected)-1) + .05, 
                label = paste("Critical Value =", round (critical_val,2))), angle = 90)+
  geom_text(aes(x = chi_square+2, y = dchisq(critical_val, length(or_expected)-1) + .05, 
                label = paste("Critical Value =", round (chi_square,2))), angle = 90)+
  labs(x = "Chi-Square", y = "Density", title = "Area under the curve") +
  theme_light(base_size = 20)
```

---

```{r, echo = 2:3}
p.usa = (table(school_usa$Superpower)/sum(table(school_usa$Superpower)))
p.usa
chisq.test(x = or_observed, p = p.usa)
```

The Oregon student preferences are not unusual under the null hypothesis (USA preferences). 

---

```{r}
c.test = chisq.test(x = or_observed, p = p.usa)
str(c.test)
c.test$residuals
```

---

```{r, echo = 2}
p.usa = as.data.frame(p.usa)[,"Freq"]
lsr::goodnessOfFitTest(x = as.factor(school$Superpower), p = p.usa)
```

(Note that this function takes the raw data, not the vector of counts.)

---

What if we had used the equal proportions null hypothesis?

```{r}
lsr::goodnessOfFitTest(x = as.factor(school$Superpower))
```

Why might this be a sensible or useful test?

---

### The usefulness of $\chi^2$

How often will you conducted a $chi^2$ goodness of fit test on raw data?

--

* (Probably) never

How often will come across $\chi^2$ tests?

--

* (Probably) a lot!

The goodness of fit test is used to statistically compare the fit of nested models. 

---

[ADD stuff about model comparison here]

---

### One-sample *t*-tests

*t*-tests were developed by William Sealy Gosset, who was a chemist studying the grains used in making beer. (He worked for Guinness.)

* Specifically, he wanted to know whether particular strains of grain made better or worse beer than the standard. 

* He developed the t-test, to test small samples of beer against a population with an unknown standard deviation.
  
  * Probably had input from Karl Pearson and Ronald Fisher
  
* Published this as "Student" because Guinness wouldn't let him share company secrets. Open science for the win!

---

### One-sample t-tests

We've already been covering one-sample t-tests, but let's formally walk through some of the steps and how this differs from a z-test.

| | Z-test | t-test |
| --|--------|--------|
| $\mu$ | known | known |
|$\sigma$ | known |unknown |
| sem or $\sigma_M$ | $\frac{\sigma}{\sqrt{N}}$ | $\frac{\hat{\sigma}}{\sqrt{N}}$ |
|Probability distribution | standard normal | t |
| DF | none | N-1 |
Tails | One or two | One or two |
| Critical value ($\alpha = .05, two-tailed$ | 1.96 | Depends on DF$ |

(As a reminder, $\hat{\sigma} = s$ .)

---

### When you assume...

...you can run a parametric statistical test!

**Assumptions of the one-sample t-test**

**Normality.** We assume the sampling distribution of the mean is normally distributed. Under what two conditions can we be assured that this is true?

**Independence.** Observations in the dataset are not associated with one another. Put another way, collecting a score from Participant A doesn't tell me anything about what Participant B will say. How can we be safe in this assumption?

---

### A brief example

Using the same Census at School data, we find that Oregon students who participated in a memory game ( $N = 227$ ) completed the game in an average time of 49.1 seconds ( $s = 13.4$ ). We know that the average US student completed the game in 45.04 seconds. How do our students compare?

--

**Hypotheses**

$H_0: \mu = 45.05$

$H_1: \mu \neq 45.05$

---

.pull-left[
$$\mu = 45.05$$

$$N = 227$$

$$ \bar{X} = 49.1 $$

$$ s = 13.4 $$

]

--

.pull.right[
```{r}
t.test(x = school$Score_in_memory_game, mu = 45.05, alternative = "two.sided")
```

]

---

```{r}
lsr::oneSampleTTest(x = school$Score_in_memory_game, mu = 45.05, one.sided = FALSE)
```

