---
title: "Lab 2: R Basics and Descriptives"
output: 
  html_document: 
    theme: cosmo
    toc: yes
    toc_depth: 3
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# suppress scientific notation
options(scipen = 999)

# load libraries
library(tidyverse)
library(here)
library(rio)
library(psych)
```


# Purpose

The purpose of today's lab is to introduce you to some simple tools that will allow you to calculate and visualize *descriptive statistics* in R. This will not be a deep dive into the theoretical meaning behind descriptives, but rather a guide to some practical steps toward getting the most basic summary information out of a given dataset. Along the way, we will implement some of the skills we learned in [last week's lab](https://uopsych.github.io/psy611/labs/lab-1.html){target="_blank"}, such as creating variables, working with data frames, installing packages, and learning how to use new functions. 

Today's lab will cover:

1. [Distributions](#dist)
2. [Measures of central tendency](#ct)
3. [Measures of variability]
4. [Skew and kurtosis]
5. [Summarizing a dataset]
6. [Standardized scores]
7. [Bivariate descriptives]

We'll use the same 2015 World Happiness Report dataset from class, which you can [download here](https://drive.google.com/uc?export=download&id=1XXsB4L4wDxGkcc-JuVnotKWCnsDUsskW). If you'd like to follow along, go ahead and import the data using the `rio` package (see [here](https://uopsych.github.io/psy611/labs/lab-1.html#importing_data_into_r){target="_blank"} for a refresher on importing data.)

```{r echo=FALSE}
world_happiness = import(here("/labs/resources/lab2/purpose/world_happiness_2015.csv"))
```


***

# Distributions {#dist}


***

# Central Tendency {#ct}

For a given variable in our dataset, we often want to get a sense

## Mean

## Median

## Mode

***

# Variability

## Range

## IQR

## Variance

## Standard deviation

***

# Skew & Kurtosis

*** 

# Summarizing data

***

# Standardization

***

# Bivariate descriptives

## Covariance 

## Correlation

***

# Minihacks

## Minihack 1:

## Minihack 2:

## Minihack 3:

## Minihack 4:
