---
title: "Lab 3: Matrix Algebra"
output: 
  html_document: 
    theme: cosmo
    toc: yes
    toc_depth: 3
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# suppress scientific notation
options(scipen = 999)

# load libraries

```


# Purpose

The purpose of today's lab is to learn how to create matrices in R and how to perform basic mathematical functions on matrices using R functions. We will also focus on how you can calculate basic descriptive statistics (which we discussed in [last week's lab](https://uopsych.github.io/psy611/labs/lab-2.html){target="_blank"}) using matrix algebra. 

Today's lab will cover:

1. [Review of matrix types](#types)
1. [How to create a matrix](#create)
1. [Testing for equality](#equality)
1. [Transpose](#transpose)
1. [Addition/subtraction](#add_sub)
1. [Multiplication](#mult)
1. 


***

# Review of matrix types {#types}

__Notation and Terminology:__

$a_{rc}$ refers to the entry at row $r$ and column $c$ in matrix $a$.  
"Order" refers to the dimensions of a matrix: the number of rows and the number of columns $(r,c)$.

1. Rectangular: $r \neq c$  
\[ A = \begin{bmatrix}
1 & 5 & 7 \\
2 & 1 & 2
\end{bmatrix} \]

2. Square:  $r = c$  
\[ a = \begin{bmatrix}
2 & 9\\
4 & 6
\end{bmatrix} \]

3. Vector:  A matrix where the row or column (not both) is 1.  
\[ A = \begin{bmatrix}
5 & 6 & 9
\end{bmatrix} \]

4. Diagonal:  A square matrix where all of the elements equal zero except for those making up the principal diagonal.  
\[ A = \begin{bmatrix}
1 & 0 & 0 \\
0 & 4 & 0 \\
0 & 0 & 3
\end{bmatrix} \]

5. Identity:  The diagonal matrix with 1s along the principal diagonal.  
\[ A = \begin{bmatrix}
1 & 0 & 0 & 0 \\
0 & 1 & 0 & 0 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1
\end{bmatrix} \]

6. Null:  A matrix that consists entirely of 0s.  
\[ 0 = \begin{bmatrix}
0 & 0 & 0 \\
0 & 0 & 0
\end{bmatrix} \]

***

# How to create a matrix {#create}

We create matrices in R by using the `matrix()` function. To specify a matrix using the `matrix()` function, we need to provide the `data` (i.e., the elements, or numbers in the matrix), the number of rows (`nrow`), the number of columns (`ncol`), and then tell it whether the order of elements is entered `byrow` or not. If you're entering it row-wise, you want to set `byrow = TRUE`. If you're entering it column wise, you want to set `byrow = FALSE`.

Let's practice creating a few matrices:

```{r equality_1_spec_marices}
a_mat <- matrix(data = c(1, 2, 3, 
                         1, 2, 3), 
                nrow = 2, ncol = 3, byrow = TRUE) # spacing is irrelevant
                                                  # but I find it easier to read
print(a_mat)

b_mat <- matrix(data = c(1, 1, 
                         2, 2, 
                         3, 3), 
                nrow = 3, ncol = 2, byrow = TRUE) 
print(b_mat)

c_mat <- matrix(data = c(1, 1, 
                         2, 4, 
                         3, 5), 
                nrow = 3, ncol = 2, byrow = TRUE) 
print(c_mat)

d_mat <- matrix(data = c(1, 1,
                         2, 2, 
                         3, 3), 
                nrow = 3, ncol = 2, byrow = TRUE) 
print(d_mat)
```

*** 

# Testing for equality {#equality}

Next, we'll Check if the order is the same using a combination of `dim()`, which returns the dimensions of the matrix (rows, columns) and the equality test `==`. REMEMBER: `=` is to define things (arguments in functions & objects) and `==` tests for equality. 

We get two results for each test, which correspond to (in order): 
<br /> 
1. same number of rows? 
<br /> 
2. same number of columns?

First let's check A against the others

```{r equality_2_a}
# check a against the others
dim(a_mat) == dim(b_mat)
dim(a_mat) == dim(c_mat)
dim(a_mat) == dim(d_mat)
```

As we knew (from looking at them ourselves), a is not the same shape as b, c, and d.

Next we'll check B angainst C and D:

```{r equality_2_b}
# Check B against C and D.
dim(b_mat) == dim(c_mat)
dim(b_mat) == dim(d_mat)
```

Dimensions of b, c, and d match.

Q: Why didn't we test c and d?
A:

The dimensions of B, C, and D are the same, so let's test the equality of the elements.
```{r equality_2_c}
b_mat == c_mat
b_mat == d_mat
```

# Transpose {#transpose}

Transposing a matrix consists of exchanging the row and column for each element in the matrix.

__Example:__

\[ A = \begin{bmatrix}
1 & 2 \\
5 & 1 \\
7 & 2
\end{bmatrix} = \begin{bmatrix}
1_{11} & 2_{12} \\
5_{21} & 1_{22} \\
7_{31} & 2_{32}
\end{bmatrix} \]

To create the transpose, the following happens:

$1_{11} \rightarrow 1_{11}$

$5_{21} \rightarrow 5_{12}$

$7_{31} \rightarrow 7_{13}$

$2_{12} \rightarrow 2_{21}$

$1_{22} \rightarrow 1_{22}$

$2_{32} \rightarrow 2_{23}$

Since the order of the above matrix is 3x2, what will the order of its transpose be?  
\[ A' = \begin{bmatrix}
1 & 5 & 7 \\
2 & 1 & 2
\end{bmatrix} \]

## Transpose in R

To transpose in R, just use the function `t()`. Let's take a look at the transpose of matrix C

```{r transpose}
print(c_mat)
t(c_mat)
```

# Addition/Subtraction 

```{r}
A <- matrix(data = c(6,1, 2,10), nrow = 2, ncol = 2, byrow = T) 
B <- matrix(data = c(2,1, 1,6), nrow = 2, ncol = 2, byrow = T) 
C <- matrix(data = c(4,1, 3,2), nrow = 2, ncol = 2, byrow = T) 

# YOUR TURN: Do these problems
# (1) A + B
A + B
# (2) Is this true: A + B = B + A
A+B==B+A
# (3) Is this true: (A + B) + C = A + (B + C)
(A + B) + C == A + (B + C)
# (4) A - B
A-B
# (5) Is this true: A - B = B - A
 A - B == B - A
# (6) Is this true: (A - B) - C = A - (B - C)
 (A - B) - C == A - (B - C)
```

# Multiplication {#mult}

Two matrices are "conformable for multiplication" if they have dimensions allowing them to be multiplied (i.e. if the inner dimensions match).

## Example 1:
\[ A=\begin{bmatrix}
1 & 3 \\
2 & 1 \\
4 & 6
\end{bmatrix},
B=\begin{bmatrix}
1 & 5 & 8 & 3 \\
4 & 2 & 6 & 4 
\end{bmatrix}\]

1. What is the order of $A$? And what about $B$?
2. Is $AB$ conformable? 
3. Is $BA$ conformable? 
4. For $AB$, what will the order of the resulting matrix be?

\[AB=\begin{bmatrix}
(1 \times 1)+(3 \times 4) & (1 \times 5)+(3 \times 2) & (1 \times 8)+(3 \times 6) & (1 \times 3)+(3 \times 4) \\
(2 \times 1)+(1 \times 4) & (2 \times 5)+(1 \times 2) & (2 \times 8)+(1 \times 6) & (2 \times 3)+(1 \times 4) \\
(4 \times 1)+(6 \times 4) & (4 \times 5)+(6 \times 2) & (4 \times 8)+(6 \times 6) & (4 \times 3)+(6 \times 4)
\end{bmatrix}
\]

\[AB=\begin{bmatrix}
13 & 11 & 26 & 15 \\
6 & 12 & 22 & 10 \\
28 & 32 & 68 & 36
\end{bmatrix}
\]

## Example 2:
\[ A=\begin{bmatrix}
1 & 2 \\
3 & 2 \\
4 & 1
\end{bmatrix},
B=\begin{bmatrix}
2 & 3 & 4 \\
1 & 2 & 4 
\end{bmatrix}\]

1. What is the order of A? And what about B?
2. Is $AB$ conformable?
3. is $BA$ conformable?
4. Does $AB = BA$?  In other words, does the matrix product possess the commutative property?

*Note:* the matrix product does possess the associative property: $A(BC)=(AB)C$.


## Matrix multiplication in R

To multiple matrices in R, we have to use the matrix multiplication operator, `%*%` (potentially helpful mneumonic: it sort of looks like an M).
```{r multiplication}
a_mat <- matrix(data = c(1, 2, 
                     3, 2, 
                     4, 1), 
            nrow = 3, ncol = 2, byrow = TRUE) 
b_mat <- matrix(data = c(2, 3, 4, 
                     1, 2, 4), 
            nrow = 2, ncol = 3, byrow = TRUE)
a_mat %*% b_mat

```

## Multiplying by the Identity Matrix
### Example
\[ C=\begin{bmatrix}
2 & 5 \\
4 & 3
\end{bmatrix}
\]

1. If C is multiplied by its identity matrix ($I_{2}$), what will the resulting matrix be?
2. Does $CI_{2} = I_{2}C$?

*Note:* for your homework, you can't just say $CI_{2} = C$, you need to go through the steps of the multiplication to prove it.

### Identity Matrix in R

The simplest way to get an identity matrix in R is to use `diag(x)` where x is the the order of the matrix (the number of rows or columns).

First we'll create a matrix C.
```{r identity_1}
c_mat <- matrix(data = c(2, 5,
                         4, 3), 
                nrow = 2, ncol = 2, byrow = TRUE)
print(c_mat)
```

And next we can get the identity matrix for C, or $I_2$
```{r identity_2}
id_mat <- diag(nrow(c_mat))
# Equivalently, 
#id_mat <- diag(2)
```

And finally, multiply them together with `%*%`
```{r}
c_mat %*% id_mat
```

And we could test if $CI_2 = I_2C$ 
```{r}
c_mat %*% id_mat == id_mat %*% c_mat
```


# Minihacks {#minihacks}

Feel free to work in groups of 2-3, and ask for help when you get stuck!

## Minihack 1:

## Minihack 2: Reproducing a covariance matrix

-multiply a matrix of deviation scores by the transpose of that matrix —this gives you sum of squares 

-then multiply by a scalar 1/(n-1) to generate a covariance matrix 

-use diag() to get the variances 

-double check your answer with cov() and var() functions 

## Minihack 3: Descriptive statistics with matrices

You run an experiment to test the effectiveness of an memory-improvement intervention. As part of the study, participants take a difficult memory test at time 1, then participate in the intervention, and then retake the memory test one week later. Half the participants are assigned to a control condition, and half are assigned to the intervention. Use the following code to create a matrix representing a dataset of scores, $X_{10,3}$:

```{r create_data, eval = FALSE}
X = matrix(c(45, 52, 54, 52, 50, 72, 43, 56, 62, 47,
             41, 51, 57, 52, 45, 83, 55, 70, 75, 57,
             1, 1, 1, 1, 1, 2, 2, 2, 2, 2), 
           ncol = 3, byrow = F, 
           dimnames = list(NULL, c("T1", "T2", "G")))
```

You should how have a matrix with three columns, named T1 (memory score at time 1), T2 (memory score at time 2), and G (group, 1 = control, 2 = intervention).

Use matrix algebra to:

1. Create a new vector representing the difference between T1 and T2.

1. Calculate the average T1 score for the whole sample.

1. Calculate the average T1 score for each group.

1. Calculate the difference in T1 and T2 for each group.



