---
title: "Lab 2: R Basics and Descriptives"
output: 
  html_document: 
    theme: cosmo
    toc: yes
    toc_depth: 3
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# suppress scientific notation
options(scipen = 999)

# load libraries
library(tidyverse)
library(here)
library(rio)
library(psych)
```


# Purpose

The purpose of today's lab is to introduce you to some simple tools that will allow you to calculate and visualize **descriptive statistics** in R. This will not be a deep dive into the theoretical meaning behind descriptives, but rather a guide to some practical steps toward getting the most basic summary information out of a given dataset. Along the way, we will implement some of the skills we learned in [last week's lab](https://uopsych.github.io/psy611/labs/lab-1.html){target="_blank"}, such as creating variables, working with data frames, installing packages, and learning how to use new functions. 

Today's lab will cover:

1. [Distributions](#dist)
2. [Basic Descriptives](#basic)
3. [Summarizing a dataset](#summarize)
4. [Correlation](#corr)

***

# Getting Started

We'll use the same 2015 World Happiness Report dataset from class, which you can [download here](https://drive.google.com/uc?export=download&id=1XXsB4L4wDxGkcc-JuVnotKWCnsDUsskW). If you'd like to follow along, go ahead and import the data using the `rio` package (see [here](https://uopsych.github.io/psy611/labs/lab-1.html#importing_data_into_r){target="_blank"} for a refresher on importing data.)

If the file is saved in your `Downloads` folder, your code might look something like this:
```{r, eval = FALSE}
library(rio) # use install.packages("rio") if the 'rio' package is not already installed
world_happiness <- import("~/Downloads/world_happiness_2015.csv")
```


```{r, echo=FALSE}
world_happiness = import(here("/labs/resources/lab2/purpose/world_happiness_2015.csv"))
```

We can take a peek at the data using the `str()` function, which shows us the structure of the dataset. This tells us we have a data frame with 136 observations of 8 different variables: `Country`, `Happiness`, `GDP`, `Support`, `Life`, `Freedom`, `Generosity` and `Corruption`. 

```{r}
str(world_happiness)
```

We can also peek at the first few rows of the data frame using `head()`.

```{r}
head(world_happiness)
```

***

# Distributions {#dist}

Recall from lecture that a **distribution** often refers to a description of the (relative) number of times a given variable will take each of its unique values.

## Histogram

One common way of visualizing distributions is using a **histogram**, which plots the frequencies of different values for a given variable.

For example, let's take a look at a distribution of the `Happiness` variable. We do this using the `hist()` function. (Remember, you can check out the help documentation using `?hist`).

```{r}
hist(x = world_happiness$Happiness)
```

This looks ok, but it's kind of ugly. Let's make it better by adding our own title and a better label for the x-axis. We can also change the number of bars using the `breaks` argument. 

```{r}
hist(x = world_happiness$Happiness, breaks = 30, main = "Histogram of Happiness Scores", xlab = "Happiness")
```

## Box Plot

We can also visualize distributions using a **boxplot**, which gives us different information. For a short guide on how to read boxplots, see [here](https://flowingdata.com/2008/02/15/how-to-read-and-use-a-box-and-whisker-plot/){target="_blank"}.

```{r}
boxplot(x = world_happiness$Happiness, main = "Boxplot of Happiness Scores")
```

## Looking ahead...

So far we have been plotting in base R. However, the [ggplot2 package](https://www.rdocumentation.org/packages/ggplot2/versions/3.2.1){target="_blank"} is generally a much better tool for plotting. For now we'll stick with base plotting to keep things simple, but in a future class you will learn how to use `ggplot` to make better-looking plots, such as this:

```{r echo = FALSE}
ggplot(world_happiness, aes(x = Happiness)) +
  geom_histogram(fill = "#56B4E9",
                 color = "white",
                 alpha = 0.7,
                 bins = 30) + 
  labs(x = "Happiness",
       y = "Frequency",
       title = "Histogram of Happiness Scores",
       caption = "Source: 2015 World Happiness Report") +
  theme_minimal(base_size = 13)
```


Ok, so now that we know how to vizualize a basic distribution, let's think about how we commonly *characterize* distributions with descriptive statistics...

***

# Basic Descriptives {#basic}

This section will describe how to calculate some of the most commonly used descriptive statistics using simple functions in R.

## Measures of Central Tendency

For a given set of observations, measures of central tendency allow us to get the "gist" of the data--e.g. they tell us about where the "average" or the "mid-point" of the data lies. Let's take a really simple example first. We'll use a vector of integers as our dataset. 

```{r}
simple_vector <- c(1, 1, 1, 2, 2, 2, 2, 2, 3, 5, 5, 6, 7)
```

### Mean

The mean is the arithmetic average of a set of values. It represents the sum of those values divided by the total number of values. Let's do this using our simple example dataset.

```{r}
sum(simple_vector)/length(simple_vector)
```

However, a faster (and better) way to find the mean is to use the aptly named `mean()` function from base R.

```{r}
mean(simple_vector)
```

What happens if we try to calculate the mean when there are missing values in our data? Let's see what happens when we add a couple `NA`'s to our simple vector and calculate the mean.

```{r}
# add NA's
simple_vector_missing <- c(1, 1, NA, 1, 2, 2, 2, 2, 2, 3, 5, NA, 5, 6, 7)

mean(simple_vector_missing)
```

...It gives us an `NA`! The reason for this is that the mean, by its nature, involves using every value for a given variable, so if you don't remove (or impute) the missing values before getting the mean, it won't work. If there are missing values in your data (which there often are in real life), then you will have to add an additional argument: `na.rm = TRUE`.

```{r}
mean(simple_vector_missing, na.rm = TRUE)
```


### Median

The median is the middle value of a set of observations: 50% of the data points fall below the median, and 50% fall above. To find the median, we can use the `median()` function from the `{stats}` package.

```{r}
median(simple_vector)
```

### Mode

Oddly enough, the core packages in R do not contain a function to find the mode. In the case of our very simple dataset, it is easy to determine that the mode (i.e. the most common value) is `2`. Your first [minihack exercise](#minihacks) will be to find the mode using the R package that accompanies your textbook.

## Measures of Variability

### Range

The range gives us the distance between the smallest and largest value in a dataset. You can find the range using the `range()` function, which will output the minimum and maximum values. 

```{r}
range(simple_vector)
```


### Variance and standard deviation

To find the variance and standard deviation, we use `var()` and `sd()`, respectively.

```{r}
# variance
var(simple_vector)

# standard deviation
sd(simple_vector)
```

Note: As it pertains to missing data, note that the same rule described above for the `mean()` function also applies to these functions (namely, that you need to add the argument `na.rm = TRUE`.)

## Outliers

Outliers are extreme values in the data. They don't follow the same general pattern as all the other data. One easy way to detect outliers is to visually inspect a distribution for values that stand out. For example, if we add an extreme value to our simple vector, it is easy to see that it is an outlier when we look at the boxplot.

```{r}
# add extreme value (30)
simple_vector_outlier <- c(1, 1, 1, 2, 2, 2, 2, 2, 3, 5, 5, 6, 7, 25)

# visualize outlier with boxplot 
boxplot(simple_vector_outlier)
```
 
### z-scores

Another way to identify outliers is to **standardize** the scores in our dataset, i.e. to convert them to z-scores. To do this, we take each score's distance from the mean and divide it by the standard deviation: 

$$ z = \frac{x_i - M}{s} $$

This tells us, for each score, how many standard deviations it is away from the mean. Some people use the rule of thumb that if a score is above or below 3 standard deviations from the mean, it should be considered an outlier. 

To standardize scores in our dataset, we can use the `scale()` function. 

```{r}
simple_vector_outlier

scale(simple_vector_outlier)
```

Based on this output, how many standard deviations above the mean is our extreme value, `25`? Should it be considered an outlier?

*** 

# Summarizing data {#summarize}

So far we have been calculating various descriptive statistics (somewhat painstakingly) using an assortment of different functions. So what if we have a dataset with a bunch of varaibles we want descriptive statistics for? Surely we don't want to calculate descriptives for each variable by hand...

Fortunately for us, there is a function called `describe()` from the `{psych}` package, which we can use to quickly summarize a whole set of variables in a dataset. Let's look back for a minute at our `world_happiness` dataset. 

## `psych::describe()`

This function automatically calculates all of the descriptives we reviewed above (and more!) 

```{r warning=FALSE}
psych::describe(world_happiness)
```

If we want to focus on a particular variable, we can calculate descriptives for that variable only.

```{r}
psych::describe(world_happiness$Happiness)
```

Note that `psych::describe()` gives us a couple statistics we haven't discussed thus far: **skew** and **kurtosis**. 

***

# Correlation {#corr}

***

# Minihacks {#minihacks}

## Minihack 1: Find the mode

1. Install and load the `{lsr}` package, the companion R package to your textbook, [Learning Statistics with R](https://learningstatisticswithr-bookdown.netlify.com/){target="_blank"}.

2. Check out the help documentation for the function `modeOf()`. 

3. Use `modeOf()` to find the mode of `simple_vector`. Make sure that the answer you get is `2`. Save your result to an object called `mode`. If you don't alreay have `simple_vector` loaded in your workspace, here it is again:

```{r eval=FALSE}
simple_vector <- c(1, 1, 1, 2, 2, 2, 2, 2, 3, 5, 5, 6, 7)
```

4. Use the function `maxFreq()` to find the frequency of the modal value (i.e. how many time it occurs). Save your result to an object called `modal_freq`.


## Minihack 2: Outliers

## Minihack 3: Descriptive plots 

## Minihack 4: Correlations
