---
title: "Lab 3: Matrix Algebra"
output: 
  html_document: 
    theme: cosmo
    toc: yes
    toc_depth: 3
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# suppress scientific notation
options(scipen = 999)

# load libraries

```


# Purpose

The purpose of today's lab is to learn how to create matrices in R and how to perform basic mathematical functions on matrices using R functions. We will also focus on how you can calculate basic descriptive statistics (which we discussed in [last week's lab](https://uopsych.github.io/psy611/labs/lab-2.html){target="_blank"}) using matrix algebra. 

Today's lab will cover:

1. [Review of matrix types](#types)
1. [How to create a matrix](#create)
1. [Testing for equality](#equality)
1. [Transpose](#transpose)
1. [Addition/subtraction](#add_sub)
1. [Multiplication](#mult)
1. 


***

# Review of matrix types {#types}

__Notation and Terminology:__

$a_{rc}$ refers to the entry at row $r$ and column $c$ in matrix $a$.  
"Order" refers to the dimensions of a matrix: the number of rows and the number of columns $(r,c)$.

1. Rectangular: $r \neq c$  
\[ A = \begin{bmatrix}
1 & 5 & 7 \\
2 & 1 & 2
\end{bmatrix} \]

2. Square:  $r = c$  
\[ a = \begin{bmatrix}
2 & 9\\
4 & 6
\end{bmatrix} \]

3. Vector:  A matrix where the row or column (not both) is 1.  
\[ A = \begin{bmatrix}
5 & 6 & 9
\end{bmatrix} \]

4. Diagonal:  A square matrix where all of the elements equal zero except for those making up the principal diagonal.  
\[ A = \begin{bmatrix}
1 & 0 & 0 \\
0 & 4 & 0 \\
0 & 0 & 3
\end{bmatrix} \]

5. Identity:  The diagonal matrix with 1s along the principal diagonal.  
\[ A = \begin{bmatrix}
1 & 0 & 0 & 0 \\
0 & 1 & 0 & 0 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1
\end{bmatrix} \]

6. Null:  A matrix that consists entirely of 0s.  
\[ 0 = \begin{bmatrix}
0 & 0 & 0 \\
0 & 0 & 0
\end{bmatrix} \]

***

# How to create a matrix {#create}

We create matrices in R by using the `matrix()` function. To specify a matrix using the `matrix()` function, we need to provide the `data` (i.e., the elements, or numbers in the matrix), the number of rows (`nrow`), the number of columns (`ncol`), and then tell it whether the order of elements is entered `byrow` or not. If you're entering it row-wise, you want to set `byrow = TRUE`. If you're entering it column wise, you want to set `byrow = FALSE`.

Let's practice creating a few matrices:

```{r equality_1_spec_marices}
a_mat <- matrix(data = c(1, 2, 3, 
                         1, 2, 3), 
                nrow = 2, ncol = 3, byrow = TRUE) # spacing is irrelevant
                                                  # but I find it easier to read
print(a_mat)

b_mat <- matrix(data = c(1, 1, 
                         2, 2, 
                         3, 3), 
                nrow = 3, ncol = 2, byrow = TRUE) 
print(b_mat)

c_mat <- matrix(data = c(1, 1, 
                         2, 4, 
                         3, 5), 
                nrow = 3, ncol = 2, byrow = TRUE) 
print(c_mat)

d_mat <- matrix(data = c(1, 1,
                         2, 2, 
                         3, 3), 
                nrow = 3, ncol = 2, byrow = TRUE) 
print(d_mat)
```

*** 

# Testing for equality {#equality}

Next, we'll Check if the order is the same using a combination of `dim()`, which returns the dimensions of the matrix (rows, columns) and the equality test `==`. REMEMBER: `=` is to define things (arguments in functions & objects) and `==` tests for equality. 

We get two results for each test, which correspond to (in order): 
<br /> 
1. same number of rows? 
<br /> 
2. same number of columns?

First let's check A against the others

```{r equality_2_a}
# check a against the others
dim(a_mat) == dim(b_mat)
dim(a_mat) == dim(c_mat)
dim(a_mat) == dim(d_mat)
```

As we knew (from looking at them ourselves), a is not the same shape as b, c, and d.

Next we'll check B angainst C and D:

```{r equality_2_b}
# Check B against C and D.
dim(b_mat) == dim(c_mat)
dim(b_mat) == dim(d_mat)
```

Dimensions of b, c, and d match.

Q: Why didn't we test c and d?
A:

The dimensions of B, C, and D are the same, so let's test the equality of the elements.
```{r equality_2_c}
b_mat == c_mat
b_mat == d_mat
```

# Transpose {#transpose}

As Elliot discussed, transposing a matrix consists of exchanging the row and column for each element in the matrix.

__Example:__

\[ A = \begin{bmatrix}
1 & 2 \\
5 & 1 \\
7 & 2
\end{bmatrix} = \begin{bmatrix}
1_{11} & 2_{12} \\
5_{21} & 1_{22} \\
7_{31} & 2_{32}
\end{bmatrix} \]

To create the transpose, the following happens:

$1_{11} \rightarrow 1_{11}$

$5_{21} \rightarrow 5_{12}$

$7_{31} \rightarrow 7_{13}$

$2_{12} \rightarrow 2_{21}$

$1_{22} \rightarrow 1_{22}$

$2_{32} \rightarrow 2_{23}$

Since the order of the above matrix is 3x2, what will the order of its transpose be?  
\[ A' = \begin{bmatrix}
1 & 5 & 7 \\
2 & 1 & 2
\end{bmatrix} \]

## Transpose in R

To transpose in R, just use the function `t()`. Let's take a look at the transpose of matrix C

```{r transpose}
print(c_mat)
t(c_mat)
```

# Addition/Subtraction 


### (4) Adding and subtracting
```{r}
A <- matrix(data = c(6,1, 2,10), nrow = 2, ncol = 2, byrow = T) 
B <- matrix(data = c(2,1, 1,6), nrow = 2, ncol = 2, byrow = T) 
C <- matrix(data = c(4,1, 3,2), nrow = 2, ncol = 2, byrow = T) 

# YOUR TURN: Do these problems
# (1) A + B
A + B
# (2) Is this true: A + B = B + A
A+B==B+A
# (3) Is this true: (A + B) + C = A + (B + C)
(A + B) + C == A + (B + C)
# (4) A - B
A-B
# (5) Is this true: A - B = B - A
 A - B == B - A
# (6) Is this true: (A - B) - C = A - (B - C)
 (A - B) - C == A - (B - C)
```


# Minihacks {#minihacks}

Feel free to work in groups of 2-3, and ask for help when you get stuck!

## Minihack 1: Find the mode {#minihack1}


