---
title: "Lab 2: R Basics and Descriptives"
output: 
  html_document: 
    theme: cosmo
    toc: yes
    toc_depth: 3
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# suppress scientific notation
options(scipen = 999)

# load libraries
library(tidyverse)
library(here)
library(rio)
library(psych)
```


# Purpose

The purpose of today's lab is to introduce you to some simple tools that will allow you to calculate and visualize *descriptive statistics* in R. This will not be a deep dive into the theoretical meaning behind descriptives, but rather a guide to some practical steps toward getting the most basic summary information out of a given dataset. Along the way, we will implement some of the skills we learned in [last week's lab](https://uopsych.github.io/psy611/labs/lab-1.html){target="_blank"}, such as creating variables, working with data frames, installing packages, and learning how to use new functions. 

Today's lab will cover:

1. Distributions (#dist)
1. Measures of central tendency (#ct)
2. Measures of variability
3. Skew and kurtosis
4. Summarizing a dataset
5. Standardized scores
6. Bivariate descriptives

We'll use the same 2015 World Happiness Report dataset from class, which you can [download here](https://drive.google.com/uc?export=download&id=1XXsB4L4wDxGkcc-JuVnotKWCnsDUsskW). 

```{r echo=FALSE}
world_happiness = import(here("/labs/resources/lab2/purpose/world_happiness_2015.csv"))
```


***

# Distributions {#dist}


***

# Central Tendency {#ct}

For a given variable in our dataset, we often want to get a sense

## Mean

## Median

## Mode



