---
title: "Lab 2: R Basics and Descriptives"
output: 
  html_document: 
    theme: cosmo
    toc: yes
    toc_depth: 3
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# suppress scientific notation
options(scipen = 999)

# load libraries
library(tidyverse)
library(here)
library(rio)
library(psych)
```


# Purpose

The purpose of today's lab is to introduce you to some simple tools that will allow you to calculate and visualize **descriptive statistics** in R. This will not be a deep dive into the theoretical meaning behind descriptives, but rather a guide to some practical steps toward getting the most basic summary information out of a given dataset. Along the way, we will implement some of the skills we learned in [last week's lab](https://uopsych.github.io/psy611/labs/lab-1.html){target="_blank"}, such as creating variables, working with data frames, installing packages, and learning how to use new functions. 

Today's lab will cover:

1. [Distributions](#dist)
2. [Basic Descriptives](#basic)
3. [Summarizing a dataset](#summarize)
4. [Bivariate descriptives](#bivar)

***

# Getting Started

We'll use the same 2015 World Happiness Report dataset from class, which you can [download here](https://drive.google.com/uc?export=download&id=1XXsB4L4wDxGkcc-JuVnotKWCnsDUsskW). If you'd like to follow along, go ahead and import the data using the `rio` package (see [here](https://uopsych.github.io/psy611/labs/lab-1.html#importing_data_into_r){target="_blank"} for a refresher on importing data.)

If the file is saved in your `Downloads` folder, your code might look something like this:
```{r, eval = FALSE}
install.packages("rio") # if the 'rio' package is not already installed
library(rio)
world_happiness <- import("~/Downloads/world_happiness_2015.csv")
```


```{r, echo=FALSE}
world_happiness = import(here("/labs/resources/lab2/purpose/world_happiness_2015.csv"))
```

We can take a peek at the data using the `str()` function, which shows us the structure of the dataset. This tells us we have a data frame with 136 observations of 8 different variables: `Country`, `Happiness`, `GDP`, `Support`, `Life`, `Freedom`, `Generosity` and `Corruption`. 

```{r}
str(world_happiness)
```

We can also peek at the first few rows of the data frame using `head()`.

```{r}
head(world_happiness)
```

***

# Distributions {#dist}

Recall from lecture that a **distribution** often refers to a description of the [relative] number of times a variable X will take each of its unique values.

We often visualize a distribution of a given variable using a histogram, which plots the frequencies of different values. 

For example, let's take a look at a distribution of the `Happiness` variable. We do this using the `hist()` function. (Remember, you can check out the help documentation using `?hist`).

```{r}
hist(x = world_happiness$Happiness)
```


This looks ok, but it's kind of ugly. Let's make it better by adding our own title and a better label for the x-axis. We can also change the number of bars using the `breaks` argument. 

```{r}
hist(x = world_happiness$Happiness, breaks = 30, main = "Distribution of Happiness Scores", xlab = "Happiness")
```

This looks a little cleaner. So far we have been plotting in base R. However, the [ggplot2 package](https://www.rdocumentation.org/packages/ggplot2/versions/3.2.1){target="_blank"} is generally a much better tool for plotting. For now we'll stick with base plotting, but in a future class you will learn how to use `ggplot` to make better-looking plots, such as this:

```{r echo = FALSE}
ggplot(world_happiness, aes(x = Happiness)) +
  geom_histogram(fill = "#56B4E9",
                 color = "white",
                 alpha = 0.7,
                 bins = 30) + 
  labs(x = "Happiness",
       y = "Frequency",
       title = "Distribution of Happiness Scores",
       caption = "Source: 2015 World Happiness Report") +
  theme_minimal(base_size = 13)
```


Ok, so now that we know how to vizualize a basic distribution, let's think about how we commonly *characterize* distributions with descriptive statistics...

***

# Basic Descriptives {#basic}

## Central Tendency

For a given set of observations, measures of central tendency allow us to get the "gist" of the data--e.g. they tell us about where the "average" or the "mid-point" of the data lies. Let's take a really simple example first...the numbers 1 through 5. 

```{r}
simple_vector <- c(1, 2, 2, 3, 7)
```

### Mean

The mean is the arithmetic average of a set of values. It represents the sum of those values divided by the total number of values. To find the mean of a set of observations, we use the aptly named `mean()` function.

```{r}
mean(simple_vector)
```


### Median

The median is the middle value of a set of observations: 50% of the data points fall below the median, and 50% fall above. To find the median, we can use `median()` from the `{stats}` package.

```{r}
median(simple_vector)
```

### Mode

Oddly enough, the core packages in R do not contain a function to find the mode. In the case of our very simple dataset, it is easy to determine that the mode (i.e. the most common value) is `2`. Your first minihack exercise will be to find the mode of a variable from a real-world dataset using a lesser-known function in R.

## Variability

### Range

### IQR

### Variance and standard deviation

## Outliers

### z-scores

*** 

# Summarizing data {#summarize}

## using `psych::describe()`

## Skew and kurtosis

***

# Standardization

***

# Bivariate descriptives {#bivar}

## Covariance 

## Correlation

***

# Minihacks

## Minihack 1:

## Minihack 2:

## Minihack 3:

## Minihack 4:
