---
title: "Lab: Introduction to R and RStudio"
output: 
  html_document: 
    theme: cosmo
    toc: yes
    toc_depth: 2
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Downloading R and RStudio

# Features of RStudio
## Source
## Console
## Enivornment/History
## Files/Plots/Packages/Help

# Packages
## Installing a package
```{r,eval=FALSE}
install.packages("yarrr")
```

## Loading a package
```{r}
library(yarrr)
```
# Try commands in R

Adapted from [__YaRrr: A Pirate's Guide to R__](https://bookdown.org/ndphillips/YaRrr/).

```{r}
?pirates

head(pirates)

mean(pirates$age)
max(pirates$age)
table(pirates$sex)

plot(pirates$height, pirates$weight)

#customize
plot(x = pirates$height, y = pirates$weight, 
     main = "Look, Ma! A scatterplot!",
     xlab = "Height (in cm)",
     ylab = "Weight (in cm)",
     pch = 16, 
     col = gray(.0, .1))

pirateplot(formula = age ~ sword.type,
           data = pirates, 
           main = "Pirateplot of age by favorite sword")

# run a t-test
t.test(formula = age ~ headband, 
       data = pirates, 
       alternative = "two.sided")

# Create tattoos model
tat.sword.lm <- lm(formula = tattoos ~ sword.type,
                   data = pirates)

# Get ANOVA table
anova(tat.sword.lm)
```

# Commands
## Arithmetic commands
## Functions
### Names
### Arguments
# Variables
## Creating
## Altering elements
# Functions and vectors
## Arithmetic functions on vectors
## `Length()`

