---
title: 'Describing Data II'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, rladies, rladies-fonts, "my-theme.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r, include = F}
knitr::opts_chunk$set(message = F, warning = F)
```


```{r, echo = FALSE, warning=FALSE, message=FALSE}
library(MASS) # for simulating variables with specific correlations
library(tidyverse) # for graphing
library(ggExtra) # for marginal distributions
library(psych) # for dataset showing content overlap
```
class: inverse

### Annoucements

- That quiz, huh?

- Homework 1 due in two weeks 

- Measure yourself for an in-class demo today [tinyurl.com/uwn463vj](https://tinyurl.com/uwn463vj)
  * height (inches)
  * length of right forearm (inches)

---

### Last time

.pull-left[### Population Variability

**Sums of squares**
$$\small SS = \Sigma(X_i-\mu_x)^2$$
**Variance**
$$\small \sigma^2 = \frac{\Sigma(X_i-\mu_x)^2}{N} = \frac{SS}{N}$$
**Standard deviation**
$$\scriptsize \sigma = \sqrt{\frac{\Sigma(X_i-\mu_x)^2}{N}}= \sqrt{\frac{SS}{N}} = \sqrt{\sigma^2}$$]


.pull-right[
###Sample variability
**Sums of squares**
$$\small SS = \Sigma(X_i-\bar{X})^2$$
**Variance**
$$\small \hat{\sigma}^2 = s^2 = \frac{\Sigma(X_i-\bar{X})^2}{N-1} = \frac{SS}{N-1}$$
**Standard deviation**
$$\scriptsize \hat{\sigma} = s = \sqrt{\frac{\Sigma(X_i-\bar{X})^2}{N-1}}= \sqrt{\frac{SS}{N-1}} = \sqrt{s^2}$$]
---

## Bi-variate descriptives

### Covariation 

"Sum of the cross-products"


### Population
$$SP_{XY} =\Sigma(X_i−\mu_X)(Y_i−\mu_Y)$$

### Sample
$$ SP_{XY} =\Sigma(X_i−\bar{X})(Y_i−\bar{Y})$$


???

**What does a large, positive SP indicate?**
A positive relationship, same sign

**What does a large, negatve SP indicate?**
A negative relationship, different sign

**What does SP close to 0 indicate?**
No relationship



---

## Covariance
Sort of like the variance of two variables

### Population
$$\sigma_{XY} =\frac{\Sigma(X_i−\mu_X)(Y_i−\mu_Y)}{N}$$

### Sample
$$s_{XY} = cov_{XY} =\frac{\Sigma(X_i−\bar{X})(Y_i−\bar{Y})}{N-1}$$


---

## Covariance table


.large[
$$\Large \mathbf{K_{XX}} = \left[\begin{array}
{rrr}
\sigma^2_X & cov_{XY} & cov_{XZ} \\
cov_{YX} & \sigma^2_Y & cov_{YZ} \\
cov_{ZX} & cov_{ZY} & \sigma^2_Z
\end{array}\right]$$
]

???
Point out that $cov_{xy}$ is the same as $cov_{yx}$

**Write on board:**

$cov_{xy} = 126.5$
$cov_{xz} = 5.2$

Which variable, Y or Z, does X have greater relationship with?
Can't know because you don't know what units they're measured in!

---

## Correlation

- Measure of association

- How much two variables are *linearly* related

- -1 to 1

- Sign indicates direction of relationship

- Invariant to changes in mean or scaling

---

## Correlation

Pearson product moment correlation

### Population
$$\rho_{XY} = \frac{\Sigma z_Xz_Y}{N} = \frac{SP}{\sqrt{SS_X}\sqrt{SS_Y}} = \frac{\sigma_{XY}}{\sigma_X \sigma_Y}$$
### Sample
$$r_{XY} = \frac{\Sigma z_Xz_Y}{n-1} = \frac{SP}{\sqrt{SS_X}\sqrt{SS_Y}} = \frac{s_{XY}}{s_X s_Y}$$
???

**Why is it called the Pearson Product Moment correlation?**
Pearson = Karl Pearson
Product = multiply
Moment = variance is the second moment of a distribution
---

```{r, echo = FALSE}
set.seed(101019) # so we all get the same random numbers
mu = c(50, 5)
Sigma = matrix(c(.8, .5, .5, .7), ncol =2) #diagonals are reliabilites, off-diagonals are correlations
data = mvrnorm(n = 150, mu = mu, Sigma = Sigma)
data = as.data.frame(data)
colnames(data) = c("x", "y")
```

```{r}
data %>% ggplot(aes(x = x, y = y)) + geom_point(size = 3) + theme_bw()
```

What is the correlation between these two variables?

???
Correlation = `r round(cor(data$x, data$y),2)`

---
```{r, echo = FALSE}
set.seed(101019) # so we all get the same random numbers
mu = c(10, 100)
Sigma = matrix(c(.8, -.3, -.3, .7), ncol =2) #diagonals are reliabilites, off-diagonals are correlations
data = mvrnorm(n = 150, mu = mu, Sigma = Sigma)
data = as.data.frame(data)
colnames(data) = c("x", "y")
```

```{r}
data %>% ggplot(aes(x = x, y = y)) + geom_point(size = 3) + theme_bw()
```

What is the correlation between these two variables?

???
Correlation = `r round(cor(data$x, data$y),2)`

---
```{r, echo = FALSE}
set.seed(101019) # so we all get the same random numbers
mu = c(3, 4)
Sigma = matrix(c(.8, 0, 0, .7), ncol =2) #diagonals are reliabilites, off-diagonals are correlations
data = mvrnorm(n = 150, mu = mu, Sigma = Sigma)
data = as.data.frame(data)
colnames(data) = c("x", "y")
```

```{r}
data %>% ggplot(aes(x = x, y = y)) + geom_point(size = 3) + theme_bw()
```

What is the correlation between these two variables?

???
Correlation = `r round(cor(data$x, data$y),2)`

---

## Effect size

- Recall that *z*-scores allow us to compare across units of measure; the products of standardized scores are themselves standardized. 

- The correlation coefficient is a **standardized effect size** which can be used communicate the strength of a relationship.

- Correlations can be compared across studies, measures, constructs, time. 

- Example: the correlation between age and height among children is $r = .70$. The correlation between self- and other-ratings of extraversion is $r = .25$. 

---

## In-class demo

Add your height (in inches), forearm length (in inches), and gender to this spreadsheet: [tinyurl.com/uwn463vj](https://tinyurl.com/uwn463vj)


???

Correlation in Pearson & Lee (1903) is about .627.

Calculate correlation in class

* Why does Pearson find smaller correlation? Students generate ideas

Show disaggregated correlation by gender (no graph)
---

## What is a large correlation?

--
- [Cohen (1988)](http://www.utstat.toronto.edu/~brunner/oldclass/378f16/readings/CohenPower.pdf): .1 (small), .3 (medium), .5 (large)
    - Often forgot: Cohen said only to use them when you had nothing else to go on, and has since regretted even suggesting benchmarks to begin with. 

--

- [Rosenthal & Rubin (1982)](https://psycnet.apa.org/fulltext/1982-22591-001.pdf): life and death (the Binomial Effect Size Display)
  - treatment success rate $= .50 + .5(r)$ and the control success rate $= .50 - .5(r)$.

--

- $r^2$: Proportion of variance "explained"
    - [Ozer & Funder (2019)](https://uopsych.github.io/psy611/readings/Ozer_Funder_2019.pdf) claim this is misleading and nonsensical
    - Fisher (2019) suggests this particular argument is non-scientific (follow up [here](https://twitter.com/aaronjfisher/status/1168252269797629952) and then [here](https://www.psychologytoday.com/us/blog/personal-and-precise/201909/can-we-get-it-right-without-needing-be-right))

---

# Funder & Ozer (2019)

- Effect sizes are often mis-interpreted. How?

- What can fix this?

- Pitfalls of small effects and large effects

- Recommendations?

???

"benchmarks" lack context and are arbitrary

fix:
- better benchmarks 
  - compare with classic studies or other literature or all literature
  - compare with non-psychological phenomenon
- explain in terms of consequences
  - BESD
  - long-run consequences

**Recommendations**
- Report effect sizes 
- Use large samples -- remember bias?
- Report effect sizes in context
- Stop using empty terminology
- Revise guildelines

---

## What affects correlations?

It's not enough to calculate a correlation between two variables. You should always look at a figure of the data to make sure the number accurately describes the relationship. Correlations can be easily fooled by qualities of your data, like:

- Skewed distributions

- Outliers

- Restriction of range

- Nonlinearity

---

## Skewed distributions

```{r, echo = FALSE}
set.seed(101019) # so we all get the same random numbers
mu = c(3, 4)
Sigma = matrix(c(.8, .2, .2, .7), ncol =2) #diagonals are reliabilites, off-diagonals are correlations
data = mvrnorm(n = 150, mu = mu, Sigma = Sigma)
data = as.data.frame(data)
colnames(data) = c("x", "y")
data$x = data$x^4
```

```{r, fig.height = 5}
p = data %>% ggplot(aes(x=x, y=y)) + geom_point()
ggMarginal(p, type = "density")
```
---

## Outliers

```{r, echo = FALSE}
set.seed(101019) # so we all get the same random numbers
mu = c(3, 4)
Sigma = matrix(c(.8, 0, 0, .7), ncol =2) #diagonals are reliabilites, off-diagonals are correlations
data = mvrnorm(n = 50, mu = mu, Sigma = Sigma)
data = as.data.frame(data)
colnames(data) = c("x", "y")
data[51, ] = c(7, 10)
```

```{r, fig.height = 5}
data %>% ggplot(aes(x=x, y=y)) + geom_point() 
```
---

## Outliers

```{r, fig.height = 5}
data %>% ggplot(aes(x=x, y=y)) + geom_point() + geom_smooth(method = "lm", se = FALSE, color = "red") + geom_smooth(data = data[-51,], method = "lm", se = FALSE)
```

---

## Outliers

```{r, echo = FALSE}
set.seed(101019) # so we all get the same random numbers
mu = c(3, 4)
n = 15
Sigma = matrix(c(.9, .8, .8, .9), ncol =2) #diagonals are reliabilites, off-diagonals are correlations
data = mvrnorm(n = n, mu = mu, Sigma = Sigma)
data = as.data.frame(data)
colnames(data) = c("x", "y")
data[n+1, ] = c(1.5, 5.5)

```

```{r, fig.height = 5}
data %>% ggplot(aes(x=x, y=y)) + geom_point() + geom_smooth(method = "lm", se = FALSE, color = "red") + geom_smooth(data = data[-c(n+1),], method = "lm", se = FALSE)
```
---

## Restriction of range

```{r, echo = FALSE}
set.seed(1010191) # so we all get the same random numbers
mu = c(100, 4)
Sigma = matrix(c(.7, .4, 4, .75), ncol = 2) #diagonals are reliabilites, off-diagonals are correlations
data = mvrnorm(n = 150, mu = mu, Sigma = Sigma)
data = as.data.frame(data)
colnames(data) = c("x", "y")
real_data = data
data = filter(data, x >100 & x < 101)
```

```{r, fig.height = 5}
data %>% ggplot(aes(x=x, y=y)) + geom_point() + geom_smooth(method = "lm", se = FALSE, color = "red")
```

???

What if I told you there were scores on X could range from 97 to 103?

---

## Restriction of range

```{r, fig.height = 5}
data %>% ggplot(aes(x=x, y=y)) + geom_point() + geom_smooth(method = "lm", se = FALSE, color = "red") + geom_point(data = real_data) + geom_smooth(method = "lm", se = FALSE, data = real_data, color = "blue")
```

???

**Can you think of example where this might occur in psychology?**
My idea: that many psychology studies only look at undergraduates (restricted age, restricted education) -- can't use these as predictors or covariates

---

## Nonlinearity

```{r, echo = FALSE, warning = FALSE}
x = runif(n = 150, min = -2, max = 2)
y = x^2 +rnorm(n = 150, sd = .5)
data = data.frame(x,y)
```

```{r, fig.height = 5}
data %>% ggplot(aes(x=x, y=y)) + geom_point() + geom_smooth(method = "lm", se = FALSE, color = "red")
```

---

## It's not always apparent

Sometimes issues that affect correlations won't appear in your graph, but you still need to know how to look for them.

- Low reliability

- Content overlap

- Multiple groups

---

## Reliability

Recall from last week:

$$r_{xy} = \rho_{xy}\sqrt{r_{xx}r_{yy}}$$

Meaning that our estimate of the population correlation coefficient is attenuated in proportion to reduction in reliability.  

**If you have a bad measure of X or Y, you will have a lower estimate of $\rho$.**

---

## Content overlap

If your Operation Y of Construct B includes items (or tasks or manipulations) that could also be influenced by Constrct A, then the correlation between X and Y will be inflated.

- Example: SAT scores and IQ tests
- Example: Depression and number of hours sleeping


- Which kind of validity is this associated with?
---
## Multiple groups

```{r, echo = FALSE}
set.seed(101019) # so we all get the same random numbers
m_mu = c(100, 4)
m_Sigma = matrix(c(.7, .4, 4, .75), ncol = 2) #diagonals are reliabilites, off-diagonals are correlations
m_data = mvrnorm(n = 150, mu = m_mu, Sigma = m_Sigma)
m_data = as.data.frame(m_data)
colnames(m_data) = c("x", "y")

f_mu = c(102, 3)
f_Sigma = matrix(c(.7, .4, 4, .75), ncol = 2) #diagonals are reliabilites, off-diagonals are correlations
f_data = mvrnorm(n = 150, mu = f_mu, Sigma = f_Sigma)
f_data = as.data.frame(f_data)
colnames(f_data) = c("x", "y")

m_data$gender = "male"
f_data$gender = "female"

data = rbind(m_data, f_data)
```

```{r, fig.height = 5}
data %>% ggplot(aes(x=x, y=y)) + geom_point() + geom_smooth(method = "lm", se = FALSE, color = "red")
```

---

## Multiple groups

```{r, fig.height = 5}
data %>% ggplot(aes(x=x, y=y, color = gender)) + geom_point() + geom_smooth(method = "lm", se = FALSE) + guides(color = "none")
```
---

### Special cases of the Pearson correlation

- **Spearman correlation coefficient**
    - Applies when both X and Y are ranks (ordinal data) instead of continuous
    - Denoted $\rho$ by your textbook, although I prefer to save Greek letters for population parameters. 

- **Point-biserial correlation coefficient**
    - Applies when Y is binary.
        - NOTE: This is not an appropriate statistic when you [artificially dichotomize data](../readings/Cohen_1983.pdf).

- **Phi (
$\phi$
) coefficient**
  - Both X and Y are dichotomous.


---
## Do the special cases matter?

For Spearman, you'll get a different answer. 
```{r}
x = rnorm(n = 10); y = rnorm(n = 10) #randomly generate 10 numbers from normal distribution
```

.pull-left[
```{r, echo = T}
head(cbind(x,y))
cor(x,y, method = "pearson")
```
]
.pull-right[
```{r}
head(cbind(x,y, rank(x), rank(y)))
cor(x,y, method = "spearman")
```
]

---
## Do the special cases matter?

If your data are naturally binary, no difference between Pearson and point-biserial.

```{r}
x = rnorm(n = 10); y = rbinom(n = 10, size = 1, prob = .3)
head(cbind(x,y))
```

.pull-left[
```{r, echo = T}
cor(x,y, method = "pearson")
```
]
.pull-right[
```{r}
ltm::biserial.cor(x,y)
```
]
---
## Do the special cases matter?

If your data are artifically binary, there can be big differences.

```{r}
x = rnorm(n = 10); y = rnorm(n = 10)
```

.pull-left[
```{r, echo = T}
head(cbind(x,y))
cor(x,y, method = "pearson")
```
]
.pull-right[
```{r}
d_y = ifelse(y < median(y), 0, 1)
head(cbind(x,y, d_y))
ltm::biserial.cor(x,d_y)
```
]

### Don't use median splits!

???

Why? 
* Median spilts can change relationship of observation to center (observation which is higher than mean may be lower than median)
* Correlations are based on moments of distribution (mean and variance), and rescaling to median violates this assumption

---

### Special cases of the Pearson correlation

Why do we have special cases of the correlation?

- Sometimes we get different results
  - If we treat ordinal data like interval/ratio data, our estimate will be incorrect
  
- Sometimes we get the same result

  - Even when formulas are different
  
  - Example: Point biserial formula
      - $$r_{pb} = \frac{M_1-M_0}{\sqrt{\frac{1}{n-1}\Sigma(X_i-\bar{X})^2}}\sqrt{\frac{n_1n_0}{n(n-1)}}$$

  
???

We don't have different formulas because the correlation is mathematically different -- these formulas were developed when we did things by hand. These formulas are short cuts!

---

class: inverse

## Next time...

latent variables (podcast!)