---
title: 'Matrix Algebra'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, rladies, rladies-fonts, "my-theme.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

## Last time

Correlations, covariances

---

## Matrix algebra

- Underlying mathematics of statistics includes a lot of algebra and calculus
  - Many textbooks and academic methods papers will use matrix algebra to explain statistical concepts

- Goal of this class: understand **why** 
   - Understand how statistics are related to each other, what factors can change your results, what's going on under the hood

--

- Another goal: find the framework for statistics that makes sense to you
  
---

## Matrix

A *matrix* is a rectangular array of numbers with n rows and m columns. It is symbolized with a bold, upper case letter, and subscripted to indicate its *order.*


 
$$\LARGE \mathbf{G_{3,2}} = \left[\begin{array}
{rrr}
g_{11} & g_{12}  \\
g_{21} & g_{22} \\
g_{31} & g_{32} 
\end{array}\right]$$



The individual elements in a matrix are called *scalars*, subscripted to indicated their position in the matrix.

---

### Vectors

$$\LARGE \mathbf{G_{3,2}} = \left[\begin{array}
{rrr}
g_{11} & g_{12}  \\
g_{21} & g_{22} \\
g_{31} & g_{32} 
\end{array}\right]$$



The columns in a matrix are called *vectors*, and are symbolized with lower case, bold letters. This matrix has two vectors 
$(\large \mathbf{g}_1$, $\large \mathbf{g}_2)$, each containing three scalars.

---

### Transpose




$$\Large \mathbf{g_{3,1}} = \left[\begin{array}
{rrr}
g_{11}\\
g_{21}\\
g_{31} 
\end{array}\right]$$



This vector is also a 3 x 1 matrix. When it is displayed as a row, it is symbolized differently:




$$\Large \mathbf{g'_{1,3}} = \left[\begin{array}
{rrr}
g_{11} & g_{12} & g_{13} \\
\end{array}\right]$$



This is called the *transpose* of the vector.


---

### Square vs rectangular

This matrix is *rectangular*. 

$$\large \mathbf{G_{3,2}} = \left[\begin{array}
{rrr}
g_{11} & g_{12}  \\
g_{21} & g_{22} \\
g_{31} & g_{32} 
\end{array}\right]$$

When the number of rows and columns are equal, the matrix is *square*:

$$\large \mathbf{F_{3,3}} = \left[\begin{array}
{rrr}
f_{11} & f_{12} & f_{13}  \\
f_{21} & f_{22} & f_{23} \\
f_{31} & f_{32} & f_{33} 
\end{array}\right]$$


Data matrices are typically rectangular; correlation matrices and covariance matrices are always square.


---

### Symmetric



If $f_{i,j} = f_{j,i}$ for all i and j, the matrix is *symmetric.*



$$\Large \mathbf{F_{3,3}} = \left[\begin{array}
{rrr}
1 & 3 & -4  \\
3 & 11 & 7 \\
-4 & 7 & 2  
\end{array}\right]$$

Correlation matrices and covariance matrices are symmetric.

---

### Diagonal



If all elements of a symmetric matrix except the main diagonal are zero, the matrix is a *diagonal* matrix:



$$\Large \mathbf{F_{3,3}} = \left[\begin{array}
{rrr}
1 & 0 & 0  \\
0 & 11 & 0 \\
0 & 0 & 2  
\end{array}\right]$$



$\mathbf{F}$ is a symmetric, diagonal matrix. 

---
## Matrix algebra

Matrix algebra shares some of the same properties as linear algebra...

... but not all of them.

Let's cover the rules of matrix algebra. 

---
##Matrix Addition and Subtraction:

- Matrices of the same order can be added and subtracted.

- These operations take place element by element.

.pull-left[
$$\large \mathbf{F_{3,2}} = \left[\begin{array}
{rrr}
1 & \underline{3}  \\
3 & 11  \\
-4 & 7   
\end{array}\right]$$
]
.pull-right[
$$\large \mathbf{H_{3,2}} = \left[\begin{array}
{rrr}
4 & \underline{-1}  \\
6 & 2  \\
12 & 8   
\end{array}\right]$$
]




$$\mathbf{F_{3,2}} + \mathbf{H_{3,2}} = \mathbf{K_{3,2}} = \left[\begin{array}
{rrr}
1+4 & \underline{3+-1}  \\
3+6 & 11+2  \\
-4+12 & 7+8   
\end{array}\right] = 
\left[\begin{array}
{rrr}
5 & \underline{2}  \\
9 & 13  \\
8 & 15   
\end{array}\right]$$

---

.pull-left[
$$\large \mathbf{F_{3,2}} = \left[\begin{array}
{rrr}
1 & \underline{3}  \\
3 & 11  \\
-4 & 7   
\end{array}\right]$$
]
.pull-right[
$$\large \mathbf{H_{3,2}} = \left[\begin{array}
{rrr}
4 & \underline{-1}  \\
6 & 2  \\
12 & 8   
\end{array}\right]$$
]




$$\mathbf{F_{3,2}} - \mathbf{H_{3,2}} = \mathbf{K_{3,2}} = \left[\begin{array}
{rrr}
1-4 & \underline{3--1}  \\
3-6 & 11-2  \\
-4-12 & 7-8   
\end{array}\right] = 
\left[\begin{array}
{rrr}
-3 & \underline{4}  \\
-3 & 9  \\
-16 & -1   
\end{array}\right]$$

???

Matrices must have the same dimensions to be added/subtracted
---

### Properties of addition and subtraction


**Matrix addition is commutative:**

$$\large \mathbf{A}+\mathbf{B}=\mathbf{B}+\mathbf{A}$$

**and associative:**

$$\large \mathbf{A} + \mathbf{B} + \mathbf{C} = (\mathbf{A} + \mathbf{B}) + \mathbf{C} = \mathbf{A} + (\mathbf{B} + \mathbf{C})$$

**Matrix subtraction is distributive:**

$$\large \mathbf{A} – (\mathbf{B} + \mathbf{C}) = \mathbf{A} – \mathbf{B} – \mathbf{C}$$ 

$$\large \mathbf{A} – (\mathbf{B} – \mathbf{C}) = \mathbf{A} – \mathbf{B} + \mathbf{C}$$
---

### Null matrix


A matrix of zeros is called a *zero matrix* or a *null matrix*. It is used in solving equations:



$$\small \left[\begin{array}
{rrr}
5 & 7  \\
4 & 2  \\
6 & 1   
\end{array}\right]
+
\left[\begin{array}
{rrr}
y_{11} & y_{12} \\
y_{21} & y_{22} \\
y_{31} & y_{32}
\end{array}\right] 
= 
\left[\begin{array}
{rrr}
7 & -1 \\
7 & 4 \\
8 & 3
\end{array}\right]$$



$$\small \left[\begin{array}
                {rrr}
                -5 & -7  \\
                -4 & -2  \\
                -6 & -1   
                \end{array}\right]
+
  \left[\begin{array}
         {rrr}
         5 & 7  \\
         4 & 2  \\
         6 & 1   
         \end{array}\right]
+
  \left[\begin{array}
         {rrr}
         y_{11} & y_{12} \\
         y_{21} & y_{22} \\
         y_{31} & y_{32}
         \end{array}\right] 
= 
    \left[\begin{array}
           {rrr}
           -5 & -7  \\
           -4 & -2  \\
           -6 & -1   
           \end{array}\right]
  +
    \left[\begin{array}
           {rrr}
           7 & -1 \\
           7 & 4 \\
           8 & 3
           \end{array}\right]$$
    
$$\small \left[\begin{array}
{rrr}
0 & 0  \\
0 & 0  \\
0 & 0   
\end{array}\right]
+
\left[\begin{array}
{rrr}
y_{11} & y_{12} \\
y_{21} & y_{22} \\
y_{31} & y_{32}
\end{array}\right] 
= 
\left[\begin{array}
{rrr}
2 & -8 \\
3 & 2 \\
2 & 2
\end{array}\right]$$

---

### Scalar multiplication

Scalar multiplication is accomplished by multiplying every element of a matrix by a constant scalar:

.pull-left[
$$\large \mathbf{F_{3,3}} = \left[\begin{array}
{rrr}
1 & 3 & -4  \\
3 & 11 & 7 \\
-4 & 7 & 2  
\end{array}\right]$$
]

.pull-right[
$$\large k = 5$$
]


$$\large k\mathbf{F} = \left[\begin{array}
{rrr}
1\times5 & 3\times5 & -4\times5  \\
3\times5 & 11\times5 & 7\times5 \\
-4\times5 & 7\times5 & 2\times5  
\end{array}\right]$$

---

### Properties of multiplication

Matrix multiplication requires attending to a few important rules:

- The order of multiplication is important.

- Matrices can only be multiplied if the number of columns of the first matrix is equal to the number of rows of the second matrix.

- The resulting matrix has an order equal to the number of rows of the first matrix and the number of columns of the second matrix.

---

$$\LARGE \mathbf{A}_{j,i} \times \mathbf{B}_{i,k} = \mathbf{C}_{j,k}$$
--
The elements of C are defined as:

- the scalar in the *j*th row and *k*th column of $\mathbf{C}$ is the sum of the scalar products of the *j*th row of $\mathbf{A}$  and the *k*th column of $\mathbf{B}$.

Or

$$\LARGE \displaystyle\sum_{n=1}^{i}{a_{j,n}\times b_{n,k}}$$
---


.pull-left[
$$\large \mathbf{A_{3,2}} = \left[\begin{array}
{rrr}
4 & -1\\
7 & 0 \\
6 & 2  
\end{array}\right]$$
]

.pull-right[

$$\large \mathbf{B_{2,3}} = \left[\begin{array}
{rrr}
1 & 7 & 0 \\
4 & 2 & 6 
\end{array}\right]$$

]



$$\large \mathbf{A}_{3,2}\times\mathbf{B}_{2,3} = \mathbf{C}_{3,3}$$

--

.pull-right[
$$\large \mathbf{C}_{3,3} = \left[\begin{array}
{rrr}
0 & 26 & -6   \\
7 & 49 & 0 \\
14 & 46 & 12
\end{array}\right]$$
]


.pull-left[
$$\large (4 \times 1) + (-1 \times 4) = 0$$
$$\large (7 \times 7) + (0 \times 2) = 49$$
]

---

### Matrix multiplication is not commutative

- It may not always be possible

  - $\Large \mathbf{A}_{2,3}\mathbf{B}_{3,5} = \mathbf{C}_{2,5}$ but $\Large \mathbf{B}_{3,5}\mathbf{A}_{2,3}$ is impossible.
  
- When possible, the results may not be equal:
  - $\Large \mathbf{A}_{1,3}\mathbf{B}_{3,1} = \mathbf{C}_{1,1}$ but $\Large \mathbf{B}_{3,1}\mathbf{A}_{1,3} = \mathbf{C}_{3,3}$
  
---

### Properties of multiplication

**Matrix multiplication is associative:**

$$\Large \mathbf{ABC} = \mathbf{(AB)C} = \mathbf{A(BC)}$$

**Matrix multiplication is distributive:** 

$$\Large \mathbf{A(B+C)} = \mathbf{AB} + \mathbf{AC}$$

**But order is important:**

$$\Large \mathbf{XA} + \mathbf{BX} \neq \mathbf{X(A + B)}$$
  
  
---
### Transpose


Every matrix has a *transpose* that is obtained by exchanging the rows and columns:

$$\large \mathbf{X} = \left[\begin{array}
{rrr}
4 & -1    \\
7 & 0  \\
6 & 2 
\end{array}\right]$$



$$\large \mathbf{X'} = \left[\begin{array}
{rrr}
4 & 7 & 6    \\
-1 & 0 & 2  \\
\end{array}\right]$$



---

###Transpose multiplication



Transposes are useful for arranging a matrix so that matrix multiplication is possible.


Example: A common statistical requirement is to generate the sums of squares and cross- products for a data matrix. If $\large \mathbf{X_{n,v}}$ is a matrix of scores, then $\large \mathbf{X_{nv}X_{nv}}$ is not possible. 



But, $\large \mathbf{X'_{vn}X_{nv}}$ can be carried out.

 - Also useful: the transpose of a product of matrices is equal to the product of the transposed matrices in reverse order
 
 $$\large (\mathbf{XY})' = \mathbf{Y'X'}$$

???

**Do students recognize what this matrix multiplication is trying to do?**

---

$\mathbf{D}$ is a matrix of deviation scores for 5 individuals on 2 variables.

.pull-left[
$$\mathbf{D} = \left[\begin{array}
{rrr}
d_{11} & d_{12}    \\
d_{21} & d_{22}    \\
d_{31} & d_{32}    \\
d_{41} & d_{42}    \\
d_{51} & d_{52}    
\end{array}\right]$$
]

.pull-right[
$$\mathbf{D'} = \left[\begin{array}
{rrr}
d_{11} & d_{21} & d_{31} & d_{41} & d_{51}   \\
d_{12} & d_{22} & d_{32} & d_{42} & d_{52}   \\
\end{array}\right]$$
]

$$\Large
\mathbf{D'D} = ??$$

---

$\mathbf{D}$ is a matrix of deviation scores for 5 individuals on 2 variables.

.pull-left[
$$\mathbf{D} = \left[\begin{array}
{rrr}
d_{11} & d_{12}    \\
d_{21} & d_{22}    \\
d_{31} & d_{32}    \\
d_{41} & d_{42}    \\
d_{51} & d_{52}    
\end{array}\right]$$
]

.pull-right[
$$\mathbf{D'} = \left[\begin{array}
{rrr}
d_{11} & d_{21} & d_{31} & d_{41} & d_{51}   \\
d_{12} & d_{22} & d_{32} & d_{42} & d_{52}   \\
\end{array}\right]$$
]

$$\Large
\mathbf{D'D} = \left[\begin{array}
{rr}
x_{11} & x_{12} \\
x_{21} & x_{22}
\end{array}\right]$$

--

$\large x_{11}$ and $\large x_{22}$ are sums of squared deviations.

--

$\large x_{12}$ and $\large x_{21}$ are sums of cross products of deviations.

--

What matrix is $\mathbf{D'D}$ one step away from?

???
Answer: variance/covariance matrix!

**How do we get there?**

Divide by $N-1$!

---

Multiplication by diagonal matrices is especially important in statistics and is used to accomplish rescaling (expanding, shrinking, standardizing).


Post-multiplication of a matrix X by a diagonal matrix $\mathbf{D}$ results in the *columns* of $\mathbf{X}$ being multiplied by the corresponding diagonal element in $\mathbf{D}$.


.pull-left[
$$\Large \mathbf{X} = \left[\begin{array}
{rr}
4 & 5 \\
7 & 6
\end{array}\right]$$
]
.pull-right[
$$\Large \mathbf{D} = \left[\begin{array}
{rr}
3 & 0 \\
0 & 4
\end{array}\right]$$
]



$$\Large \mathbf{XD} = \mathbf{Y} = \left[\begin{array}
{rr}
12 & 20 \\
21 & 24
\end{array}\right]$$



The first column of X is multiplied by 3; the second column is multiplied by 4.

---

Pre-multiplication of a matrix X by a diagonal matrix $\mathbf{D}$ results in the *rows* of $\mathbf{X}$ being multiplied by the corresponding diagonal element in $\mathbf{D}$.



.pull-left[
$$\Large \mathbf{D} = \left[\begin{array}
{rr}
3 & 0 \\
0 & 4
\end{array}\right]$$
]
.pull-right[
$$\Large \mathbf{X} = \left[\begin{array}
{rr}
4 & 5 \\
7 & 6
\end{array}\right]$$
]



$$\Large \mathbf{DX} = \mathbf{Y} = \left[\begin{array}
{rr}
12 & 15 \\
28 & 24
\end{array}\right]$$



The first row of X is multiplied by 3; the second row is multiplied by 4.

---

### Scalar multiplication (again)



Scalar multiplication is just multiplication by a diagonal matrix with a constant in the diagonal.



$$\Large \left[\begin{array}
{rr}
3 & 0 \\
0 & 3
\end{array}\right]
\left[\begin{array}
{rr}
4 & 7 \\
7 & 6
\end{array}\right]
= 
\left[\begin{array}
{rr}
12 & 21 \\
21 & 18
\end{array}\right]$$



$$\Large \left[\begin{array}
{rr}
4 & 7 \\
7 & 6
\end{array}\right]
\left[\begin{array}
{rr}
3 & 0 \\
0 & 3
\end{array}\right]
=
\left[\begin{array}
{rr}
12 & 21 \\
21 & 18
\end{array}\right]$$



---
### Identity matrix


The *identity matrix* is a diagonal matrix with ones on the main diagonal:



$$\Large \mathbf{I} = \left[\begin{array}
{rrrr}
1 & 0 & 0 & 0 & 0 \\ 
0 & 1 & 0 & 0 & 0 \\
0 & 0 & 1 & 0 & 0 \\
0 & 0 & 0 & 1 & 0 \\
0 & 0 & 0 & 0 & 1
\end{array}\right]$$



The identity matrix is often a useful target matrix in statistics.

---

### Inverse matrix

Some square matrices have an *inverse* such that $\large \mathbf{AA^{-1}} = I$. The inverse is useful in solving matrix equations:

$$\Large \mathbf{Y} = \mathbf{BR}$$

Solving for B:

$$\Large \mathbf{YR^{-1}} = \mathbf{BRR^{-1}}$$
$$\Large \mathbf{YR^{-1}} = \mathbf{B}$$


---

class: inverse

## Statistics through matrix algebra


---

## Linear combinations

When we create a **linear combination**, we simply mean we're weighting values and then adding them together. 

For example:

$$y = 5x_1 + \frac{1}{7}x_2 - 12x_3$$

- What are the weights?

--

Matrix algebra is simply representing one or more linear combinations by sorting the weights inside a matrix.

--

Linear combinations can be done to rows or to columns. 

---

## Linear combinations

Perhaps **the most important rules of matrix algebra** to remember are that :

- post-multiplication creates linear combinations of columns, and 
- pre-multiplication creates linear combinations of rows.

By applying pre- and post-multiplication to a dataframe of raw scores you can perform transformations to your raw data -- including calculating sum or weighted scores -- and aggregate your data into whatever groups or configurations are necessary.


---

##Linear combinations

If I create the following linear combinations of the original variables:

.pull-left[
$$\large \mathbf{X} = \left[\begin{array}
{rrr}
X_{11} & X_{12} & X_{13} \\
X_{21} & X_{22} & X_{23} \\
X_{31} & X_{32} & X_{33} \\
\vdots & \vdots & \vdots \\
X_{n1} & X_{n2} & X_{n3} 
\end{array}\right]$$
]


.pull-right[
$$\large \mathbf{T} = \left[\begin{array}
{rrr}
1 & -1 & 1 \\
1 & 0 & -2 \\
1 & 1 & 1
\end{array}\right]$$
]

$$\Large \mathbf{XT} = Y$$

What do the new variables represent? 

???

$\mathbf{Y}$ is a $n x 3$ matrix. 

Post-multiplcation = affecting the columns.


Each row of Y corresponds to one person.
First column is the sum of the person's values (e.g., sum all values in a scale)
Second column = 3rd-1st
Third column = (1st+3rd)-2*2nd

---

##Linear combinations

If I add a linear combinations as pre-multiplication:

.pull-left[
$$\mathbf{T} = \left[\begin{array}
{rr}
1 & 1 & 1 & \dots & 0 & 0 \\
1 & 1 & 1 & \dots & -1 & -1 \\
\end{array}\right]$$
]


.pull-right[
$$\mathbf{X} = \left[\begin{array}
{rrr}
X_{11} & X_{12} & X_{13} \\
X_{21} & X_{22} & X_{23} \\
X_{31} & X_{32} & X_{33} \\
\vdots & \vdots & \vdots \\
X_{n1} & X_{n2} & X_{n3} 
\end{array}\right]$$
]

$$\Large \mathbf{TX} = Y$$

What do the new variables represent? 

--

Post-multiplication can be used to create composites or indices. 
Pre-multiplication creates groups! (treatment vs control, young vs old...)

???

$\mathbf{Y}$ is a $2 x 3$ matrix. 

Pre-multiplication = affecting the rows


Each column of Y corresponds to one variable
First row is the mean of group 1
2nd row is difference in means of group 1 and group 2


---

## Descriptive statistics

We can calculate the mean of a vector $\mathbf{x}$ with length $n$ using linear combinations.

$$\mathbf{T} = \left[\begin{array} 
& \frac{1}{n}  & \frac{1}{n}  & \dots & \frac{1}{n}  & \frac{1}{n} \\
\end{array}\right]$$

This happens to be equivalent to the following formula:

$$\Large \bar{X} = T\mathbf{X} = 1'\mathbf{X}(1'1)^{-1} = 1'\mathbf{X}\frac{1}{n}$$

(1 refers to a square matrix of 1's)
---
  
In other words, the mean (of a variable) is a linear combination of rows/observations/people. 

$$\bar{x} = \frac{1}{n}x_1 + \frac{1}{n}x_2 + \frac{1}{n}x_3 + ... +\frac{1}{n}x_n $$
  
$$ \mathbf{T} = \left[\begin{array}
                         {rrr}
                         \frac{1}{n}  & \frac{1}{n} & \frac{1}{n}  & \dots & \frac{1}{n}  & \frac{1}{n} \\
                         \end{array}\right] $$
    
Our transformation matrices are simply the weights we use when adding values together. 

(Note: when we subtract a value, we assign it the weight of -1.)

---
    
Imagine we only wanted to know the mean of a subset of participants, perhaps the participants in some group, G.
  
  
Our linear transformation of all the scores would look like:
    
$$\bar{x_G} = \frac{1}{n_G}x_1 + \frac{1}{n_G}x_2 + \frac{1}{n_G}x_3 + \dots +0x_{n-1}+0x_n$$
Participants are given a weight equal to 1 over the sample size of the group if they are a member of that group. 

If they are not part of the group, they are given the weight 0. 

$$\mathbf{T} = \left[\begin{array}
                          {rrr}
                          \frac{1}{n_G}  & \frac{1}{n_G}  & \frac{1}{n_G}  & \dots & 0  & 0 \\
                          \end{array}\right]$$
    
    

---
    
With matrices, we can perform multiple calculations all at once. 

With one matrix, we can find the average of all scores, the average of scores in group 1, the average of scores in group 2 and the difference in average scores between the two groups:
    
$$\mathbf{T} = \left[\begin{array}
                          {rrrrrr}
                          \frac{1}{n} & \frac{1}{n} & \frac{1}{n} &\dots & \frac{1}{n} & \frac{1}{n} \\
                          \frac{1}{n_{G1}} & \frac{1}{n_{G1}} & \frac{1}{n_{G1}} & \dots & 0 & 0 \\
                          0 & 0 & 0 & \dots & \frac{1}{n_{G2}} & \frac{1}{n_{G2}} \\
                          \frac{1}{n_{G1}} & \frac{1}{n_{G1}} & \frac{1}{n_{G1}} & \dots & -\frac{1}{n_{G2}} & -\frac{1}{n_{G2}} \\
                          \end{array}\right]$$
    
    
If we pre-multiply this to a matrix X of scores, it will perform these four calculations for all variables (columns) in X. 

---
    

$$\small 
      \mathbf{T} = \left[\begin{array}
                          {rrrr}
                          \frac{1}{n} & \frac{1}{n} & \frac{1}{n} &\dots & \frac{1}{n} & \frac{1}{n} \\
                          \frac{1}{n_{G1}} & \frac{1}{n_{G1}} & \frac{1}{n_{G1}} & \dots & 0 & 0 \\
                          0 & 0 & 0 & \dots & \frac{1}{n_{G2}} & \frac{1}{n_{G2}} \\
                          \frac{1}{n_{G1}} & \frac{1}{n_{G1}} & \frac{1}{n_{G1}} & \dots & -\frac{1}{n_{G2}} & -\frac{1}{n_{G2}} \\
                          \end{array}\right]$$
                          


.pull-right[
$$\small \mathbf{X} = \left[\begin{array}
{lrr}
\text{Treatment} & 5 & 45 \\
\text{Treatment} & 3 & 34 \\
\text{Treatment} & 5 & 27 \\
\text{Control} & 8 & 32 \\
\text{Control} & 5 & 71 \\
\end{array}\right]$$
]

We can pre-multiply X by T:
  
  
```{r, echo = F}
  X = matrix(
    c(5,3,5,8,5,45,34,27,32,71),
    byrow = F, ncol = 2
  )
  
  T = matrix(
    c(rep(.2, 5),
      rep(1/3,3), rep(0,2),
      rep(0, 3), rep(.5,2),
      rep(1/3, 3), rep(-.5, 2)),
    byrow = T, ncol = 5)
```
```{r}
  T %*% X
```
---
<!-- ### Correlation matrix, $\large \mathbf{R}$ -->

<!-- A correlation is simply the covariance divided by the standard deviations of the two variables, so we can create a new formula to calculate that. -->

<!-- Let $\mathbf{D_{xx}}$ be a diagonal matrix with the diagonal value being $1/s_i$ or the inverse of the standard deviation of that variable. Then we can calculate $\mathbf{R}$ as -->

<!-- $$\large \mathbf{R} = \mathbf{D_{xx}S_{XY}D_{xx}}$$ -->

<!-- --- -->
    
## Matrix Algebra in statistics
    

For the next two (3?) terms, we'll use matrix algebra as one of the ways to represent the statistics we're calculating. 

--

Your statistical tool kit now includes
  - validity (pre-data collection)
  - descriptive statistics (information about your sample, estimates of the population)
  - matrix algebra (ways to combine raw data into statistics or indices)
  
At this point, we'll switch gears and talk about the branch of mathematics that drives statistical inference...

---

class: inverse
  
## Next time...
  
probability!
    
&nbsp;
&nbsp;
&nbsp;


### [Click here to take Quiz #2 on Canvas now!](https://canvas.uoregon.edu/courses/165689) 