---
title: 'Probability: Binomial Distribution'
output:
  xaringan::moon_reader:
    includes: 
    lib_dir: libs
    css: [default, rladies, rladies-fonts, "my-theme.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

### Annoucements

- Mid-term feedback in journals

- Homework #1 due Friday at 9 am
    - attach both .RMD and .html files to submission on Canvas

---

## Last week...

- Introduction to probability

- Jargon (elementary events, sample space, conditional, independence)

- Frequentist
  - probability == long run rate
- Bayesian
  - start with prior belief, incorporate data
  
- Back to frequentist
---

```{r, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
library(tidyverse)
library(gganimate)
knitr::opts_chunk$set(message = F, warning = F)
```

The **binomial distribution** is the theoretical probability distribution appropriate when modeling the expected outcome, X, of N trials (or event sequences) that have the following characteristics:

--
- The outcome on every trial is binary 

    - also called a **Bernoulli trial**

--

- The probability of the target outcome (usually called a “success”) is the same for all N trials 

    - “with replacement” might be necessary

--

- The trials are independent

--

- The number of trials is fixed

---



If these assumptions hold then X is a binomial random variable representing the expected number of successes over N trials, with expected success on each trial of $\theta$ .

<p>&nbsp;</p>


A common and compact way of stating the same thing is: 

<p>&nbsp;</p>


$$\Huge X \sim B(N, \theta)$$
---

The probability distribution for X is defined by the following **probability mass function**: 

$$\Large P(X|\theta,N) = \frac{N!}{X!(N-X)!}\theta^X(1-\theta)^{N-X}$$

The probability mass function tells us what to expect for any particular X in the sample space.

<p>&nbsp;</p>

All theoretical distributions have a mass function (if discrete) or a density function (if continuous). These are the defining equations that tells us the generating process for the behavior of X.

???

A common way to write the binomial mass function is to think of $\theta$ as the probability of success $(p)$ and $1-\theta$ as the probability of failure $(q)$. It becomes easier to write the function:

$$ P(X|\theta,N) = \frac{N!}{X!(N-X)!}p^X(q)^{N-X}$$
---

$$\Large P(X|\theta,N) = \frac{N!}{X!(N-X)!}\theta^X(1-\theta)^{N-X}$$

***

$\mathbf{P(X|\theta,N)}$ is a conditional probability: the probability of X given $\theta$ and $N$.

- X is the number of successful trials over N independent trials, with the probability of success on any trial equal to $\theta$.

- $\theta$ and N are parameters of the binomial distribution.

---

$$\Large P(X|\theta,N) = \frac{N!}{X!(N-X)!}\theta^X(1-\theta)^{N-X}$$

***

$\mathbf{\theta^X(1-\theta)^{N-X}}$ is the probability of any particular instance of X.  
- This is just a general form of the basic probability rule:

$$A \text{ and } B = P(A \cap B) = P(A)P(B)$$

Note that this form of the rule assumes *independent events*.

---
For example, let's examine a sequence of 5 independent rolls of a die:

`3  6  6  1  6`

--


This can be represented in binomial form. First we have to choose the value that represents "success." Here, we'll use 6. 

`Not6  6  6  Not6  6`

--

The probability of that particular sequence is then:

$$P(Not6)P(6)P(6)P(Not6)P(6)$$

--

$$P(6)^3P(Not6)^2 = (\frac{1}{6})^3(\frac{5}{6})^2 = `r round(((1/6)^3)*((5/6)^2), 4)`$$



---


$$\Large P(X|\theta,N) = \frac{N!}{X!(N-X)!}\theta^X(1-\theta)^{N-X}$$

***

But a specific sequence of independent outcomes is just one way we could have X successful trials out of N. 

- We need to know **how many possible ways** we could get X successes in N trials.

The remaining part of the equation (the combination rule from probability theory, $_XC_N$), tells us how many different ways that can happen.

$$\Large \frac{N!}{X!(N-X)!}$$

---

Returning to our dice example, how many ways are there to roll a six 3 times out of 5?

--

.pull-left[

`6  6  6  Not6 Not6`



`6  6  Not6  6 Not6`



`6  6  Not6  Not6 6`

`6  Not6  6  6  Not6`

`6  Not6  6  Not6  6`
]

.pull-right[

`6  Not6  Not6  6  6`

`Not6  6  6  6  Not6`

`Not6  6  6  Not6  6`

`Not6  6  Not6  6  6`

`Not6  Not6  6  6  6`
]


--

$$\large \frac{5!}{3!(5-3)!} = \frac{5\times4\times3\times2\times1}{3\times2\times1(2\times1)}=10$$
---
Putting the pieces together:

$$\large P(X = \text{a }6, \text{three times}|\theta_6, N= 5)\\
= \frac{N!}{X!(N-X)!}\theta^X(1-\theta)^{N-X}\\=
\frac{5!}{3!(5-3)!}(\frac{1}{6})^3(\frac{5}{6})^2\\
= (10)(.0032) \\
=.032$$

---

A note about notation:

Many texts refer to the probability of success as $p$ and the probability of not success (or failure) as $q$. 

In some ways, this makes the formula easier to understand:

$$P(X|p, N)= \frac{N!}{X!(N-X)!}p^Xq^{(N-X)}$$

---

What does the Law of Total Probability require to be true?

```{r binom-plot, echo = F}
data.frame(num = 0:5, p = dbinom(x = 0:5, size = 5, prob = 1/6), three = as.factor(c(0,0,0,1,0,0))) %>% ggplot(aes(x=num, y=p, fill = three)) + geom_bar(stat="identity") + scale_x_continuous("Number of sixes (X) in five rolls (N)", breaks=c(0:5)) +scale_y_continuous("Probability")+ guides(fill = "none") + ggtitle("Binomial Probability Distribution")
```

```{r ref.label = 'binom-plot', eval = FALSE}

```

???

Independent rolls!

---

Every probability distribution has an **expected value** distribution. 

For the binomial distribution:

$$E(X) = N\theta$$

--
Each probability distribution also has a variance. For the binomial:

$$Var(X) = N\theta(1-\theta)$$
--

Importantly, this means our mean and variance are related in the binomial distribution, because they both depend on $\theta$. How are they related?

--

If you have a discrete distribution with a small N, these estimates may not have a sensible meaning. 

Later we will use the variance to help us make statements about how confident we are with regard to the location of the mean. 

???

Expected value = most likely result of the probability function,
* the thing we would expect to happen if we have no other information than the parameters of the distribution. 
*the long run average over an infinite amount of trials or samplings

Sensible mean = number of arms example

---

.left-column[
The mean, .835, does not exist in the sample space, and rounding up to 1 and claiming that to be the most typical outcome is not quite right either. 
]

```{r ref.label = 'binom-plot', echo = FALSE}

```



---

The **probability mass (density) function** allows us to answer other questions about the sample space that might be more important, or at least realistic.
- mass = discrete
- density = continuous

--

I might want to know the value in the sample space at or below which a certain proportion of outcomes fall. This is a **percentile or quantile** question. 
  -  "Most (75%) of five dice rolls yield X or fewer 6's."

--

I might want to know the proportion of outcomes in the sample space that fall at or below a particular outcome. This is a **cumulative proportion** question.
  - "What percentage of the time will my outcome be less than 3?"

---
At or below what outcome in the sample space do .75 of the outcomes fall?

```{r echo = FALSE}
data.frame(num = 0:5, p = dbinom(x = 0:5, size = 5, prob = 1/6)) %>% ggplot(aes(x=num, y=p)) + geom_bar(stat="identity") + geom_vline(aes(xintercept = qbinom(p = .75, size = 5, prob = 1/6)), color = "red") +scale_x_continuous("Number of sixes (X) in five rolls (N)", breaks=c(0:5)) +scale_y_continuous("Probability")+ggtitle("Binomial Probability Distribution")
``` 

---
What proportion of outcomes in the sample space that fall at or below a given outcome?

.pull-left[
```{r ref.label = 'binom-plot', echo = FALSE}

```

]
.pull-right[
```{r echo = FALSE}
data.frame(num = 0:5, p = pbinom(q = 0:5, size = 5, prob = 1/6)) %>% ggplot(aes(x=num, y=p)) + geom_bar(stat="identity") + scale_x_continuous("Number of sixes (X) in five rolls (N)", breaks=c(0:5)) +scale_y_continuous("Probability")+ggtitle("Cumulative Distribution")
``` 
]

---

In R, we can calculate the cumulative probability (X or lower), using the `pbinom` function.

```{r}
# what is probability of rolling two or fewer 6's out of five rolls?
pbinom(q = 2, size = 5, prob = 1/6)
```


---

The binomial is of interest beyond describing the behavior of dice and coins.

Many practical outcomes might be best described by a binomial distribution.

For example, suppose I give a 40-item multiple choice test, with each question having 4 options.

* I am worried that students might do well by chance alone.  I would not want to pass students in the class if they were just showing up for the exams and guessing for each question.

* What are the parameters in the binomial distribution that will help me address this question?

???

$N = 40$
s
$\theta = .25$

---

```{r binom-plot2, echo = F}
limit = 30
data.frame(num = 0:limit, p = dbinom(x = 0:limit, size = 40, prob = .25)) %>% ggplot(aes(x=num, y=p)) + geom_bar(stat="identity") + scale_x_continuous("Test Score (0-40) Under a Guessing Model", breaks=seq(0,limit,2)) +scale_y_continuous("Probability")+ggtitle("Binomial Probability Distribution")
```

???
I could use this distribution to help me decide if a given student is consistent with a guessing model.

Nearly all of the outcomes expected for guessers fall below the minimum passing score (60%, D-, 24).

---

How likely is it that a guesser would score above the threshold (60%) necessary to pass the class by the most minimal standards?

$$P(24|.25, 40) + P(25|.25,40) + P(26|.25,40) + ... + P(40|.25, 40) $$
--

<!-- $$1-P(X = 23|\theta = .25, N= 40)= \\ 1-\frac{23!}{23!(40-23)!}(.25)^{23}(1-.25)^{17}$$ -->
```{r}
#Note the use of the Law of Total Probability here

1-pbinom(q = 23, size = 40, prob = .25)
```

---

Cumulatively, what proportion of guessers will fall below each score?

```{r, echo = F}
limit = 40
data.frame(num = 0:limit, p = pbinom(q = 0:limit, size = 40, prob = .25)) %>% ggplot(aes(x=num, y=p)) + geom_line() + geom_vline(aes(xintercept = 24), color = "red") + scale_x_continuous("Test Score (0-40) Under a Guessing Model", breaks=seq(0,limit,2)) +scale_y_continuous("Cumulative Proportion")+ggtitle("Cumulative Binomial Probability Distribution")
```

???

Seems safe to assume that, practically speaking, all guessers will fall below the minimally passing score.

---
###There’s always a but

But, what assumptions are we making and what consequences will they have?

--

* The outcome on every trial is binary (also called a Bernoulli trial)

--

* The probability of the target outcome (usually called a "success") is the same for all N trials ("with replacement" might be necessary)

--

* The trials are independent $P(A\cap B) = P(A|B)P(B)=P(A)P(B)$

--

* The number of trials is fixed

--

In probability and statistics, if the assumptions are wrong then inferences based on those assumptions could be wrong too, perhaps seriously so.

---

> All models are wrong, but some models are useful.  (G.E.P. Box)

We might have viable alternative models:

* **Geometric distribution:**  Used if we are interested in the number of trials required for one "success" to occur 
  * "how many times do I start my computer before it fails to start at all?"
  
---

> All models are wrong, but some models are useful.  (G.E.P. Box)

We might have viable alternative models:

* **Negative binomial distribution:** Used if we are interested in the number of successes in a series of repeated trials until a specified number of failures are seen 
  * "What is the probability that a baseball player will get his 2nd hit on his 4th at-bat?"
  * "A child won't return from trick or treating until they get 5 full-size candy bars. What is the probability that they will have to visit 34 homes to get this?"
  
---

> All models are wrong, but some models are useful.  (G.E.P. Box)

We might have viable alternative models:
* **Poisson distribution:** Used when there's not a fixed number of trials but rather a fixed interval of time. 
  * "What is the expected number of times a solider will be kicked in the head by a horse and die during this one year campaign?"

---

.left-column[

.small[As N increases, the binomial becomes more normal in appearance.

Because of the difficulties in calculating large factorials, there is a large-sample normal approximation to the binomial. The normal distribution is useful for a lot of other reasons too.
]
]


```{r, echo = F, eval = F}
data.frame(num = 0:30, p = dbinom(x = 0:30, size = 100, prob = 1/6)) %>% ggplot(aes(x=num, y=p)) + geom_bar(stat="identity") + scale_x_continuous("Number of sixes (X) in 100 rolls (N)", breaks=seq(0,30,2)) +scale_y_continuous("Probability")+ggtitle("Binomial Probability Distribution")
```

```{r, echo = F}
plot = data.frame(N = 5:50, theta = 1/6) %>%
  mutate(X = map(N, ~0:.x),
         density = map2(N, theta, ~dbinom(0:.x, size = .x, prob = .y))) %>%
  unnest(c(X, density)) 

plot %>%
  ggplot(aes(x=X, y=density, frame=N)) + 
  geom_bar(stat="identity") + 
  labs(x = "Number of sixes (X) in N rolls",
       y = "Probability",
       title = "N = {closest_state}") +
  transition_states(N,
                    transition_length = 10,
                    state_length = 2)

```


---

class: inverse

## Next time...

the normal distribution